---
title: "SET2 - SET3 - SET4 Comparison"
params:                              
  experiment_id: S2_S3_S4_COMP
  run_date: 2026-06-09
  author: "Shalise Burch"
output: html_document  
knitr:
  opts_chunk:
    echo: TRUE        
    warning: FALSE     
    message: FALSE     
---

STEP 0: Set up


# ------------------
# STEP 0: Setup
```{r libraries and variables, message=FALSE, warning=FALSE}
#UPDATED:
# 2026-06-08 14:30 SAB ------------------------------

library(here)
library(DESeq2)
library(limma)
library(PCAtools)
library(tximport)
library(pheatmap)
library(dplyr) 
library(ggplot2)
library(gtools)
library(ggsci)
library(stringr) 
library(edgeR)
library(grid)
library(data.table)
library(tidyverse)
library(writexl)
library(ashr)


exp_name <- params$experiment_id
data_path <- here("rnaseq", "01_data")
results_path <- here("rnaseq", "02_results")
rdata_path <- here("rnaseq", "03_rdata")
source(here("rnaseq", "00_scripts", "rnaseq_functions.R"))


# load metasheets and txi
# -- SET 2 -- 
SAB06_meta <- readRDS("/Users/shalise/Library/CloudStorage/OneDrive-TheUniversityofChicago/R_Studio/SAB06/rnaseq/03_rdata/SAB06_metadataFinal.rds")

SAB06_star_salmon_path <- "/Volumes/project/imoskowitz/shared/sequencing.processed/SAB06/star_salmon"

# -- SET 3 -- 
SAB01_meta <- readRDS("/Users/shalise/Library/CloudStorage/OneDrive-TheUniversityofChicago/R_Studio/SAB01/rnaseq/03_rdata/SAB01_metadataFinal.rds")

SAB01_star_salmon_path <- "/Volumes/project/imoskowitz/shared/sequencing.processed/20240412_LH00315_0161_B222TJLLT1-XRS-Pl21-IM-RS54-hg38/star_salmon"


# -- SET 4 -- 
SAB08_meta <- readRDS("/Users/shalise/Library/CloudStorage/OneDrive-TheUniversityofChicago/R_Studio/SAB08/rnaseq/03_rdata/SAB08_metadataFinal_outliers_removed.rds")

SAB08_star_salmon_path <- "/Volumes/project/imoskowitz/shared/sequencing.processed/SAB08/star_salmon"


save_dir <- file.path(results_path, "S2_S3_S4_COMP")
make_subdir(save_dir)

```


## Clean metasheet
```{r clean data}
# add seq_run to SAB01 and SAB06 to match SAB08
SAB01_meta$seq_run <- "1"
SAB06_meta$seq_run <- "1"

# add cell set identifier
SAB06_meta$cell_set <- "Set2"
SAB01_meta$cell_set <- "Set3"
SAB08_meta$cell_set <- "Set4"


# add new sampleID identifier
SAB06_meta <- SAB06_meta %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

SAB01_meta <- SAB01_meta %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

SAB08_meta <- SAB08_meta %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# Make sure rownames are new sample IDs
rownames(SAB06_meta) <- SAB06_meta$sampleID_unique
rownames(SAB01_meta) <- SAB01_meta$sampleID_unique
rownames(SAB08_meta) <- SAB08_meta$sampleID_unique


#check they match 
setdiff(colnames(SAB01_meta), colnames(SAB08_meta))
setdiff(colnames(SAB08_meta), colnames(SAB01_meta))
setdiff(colnames(SAB06_meta), colnames(SAB01_meta))
setdiff(colnames(SAB01_meta), colnames(SAB06_meta))


```



## Load txi object
```{r txi}

# load txi
SAB01_txi <- salmon_import(SAB01_star_salmon_path)
SAB08_txi <- salmon_import(SAB08_star_salmon_path)
SAB06_txi <- salmon_import(SAB06_star_salmon_path)


#check sample names 
colnames(SAB01_txi$counts)
rownames(SAB01_meta)

colnames(SAB08_txi$counts)
rownames(SAB08_meta)

colnames(SAB06_txi$counts)
rownames(SAB06_meta)
```


## clean txi names
### set 2
```{r clean txi names}

# Set3: strip SAB01 prefix, align metadata, then assign unique IDs
strip_prefix <- function(x) sub("^SAB06_", "", x)

# strip first
for (elt in intersect(names(SAB06_txi), c("counts", "abundance", "length"))) {
  colnames(SAB06_txi[[elt]]) <- strip_prefix(colnames(SAB06_txi[[elt]]))
}

# now align metadata to stripped txi column order
SAB06_meta <- SAB06_meta %>%
  dplyr::slice(match(colnames(SAB06_txi$counts), sampleID)) %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# rename all txi elements to unique IDs
for (elt in intersect(names(SAB06_txi), c("counts", "abundance", "length"))) {
  colnames(SAB06_txi[[elt]]) <- SAB06_meta$sampleID_unique
}

rownames(SAB06_meta) <- SAB06_meta$sampleID_unique

stopifnot(all(colnames(SAB06_txi$counts) == rownames(SAB06_meta)))
```

### set 3
```{r clean txi names}

# Set3: strip SAB01 prefix, align metadata, then assign unique IDs
strip_prefix <- function(x) sub("^SAB01_", "", x)

# strip first
for (elt in intersect(names(SAB01_txi), c("counts", "abundance", "length"))) {
  colnames(SAB01_txi[[elt]]) <- strip_prefix(colnames(SAB01_txi[[elt]]))
}

# now align metadata to stripped txi column order
SAB01_meta <- SAB01_meta %>%
  dplyr::slice(match(colnames(SAB01_txi$counts), sampleID)) %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# rename all txi elements to unique IDs
for (elt in intersect(names(SAB01_txi), c("counts", "abundance", "length"))) {
  colnames(SAB01_txi[[elt]]) <- SAB01_meta$sampleID_unique
}

rownames(SAB01_meta) <- SAB01_meta$sampleID_unique

stopifnot(all(colnames(SAB01_txi$counts) == rownames(SAB01_meta)))
```

### set4
```{r}

# strip SAB08 prefix
strip_prefix <- function(x) sub("^SAB08_", "", x)

for (elt in intersect(names(SAB08_txi), c("counts", "abundance", "length"))) {
  colnames(SAB08_txi[[elt]]) <- strip_prefix(colnames(SAB08_txi[[elt]]))
}

# samples shared between txi and metadata
keep_samples <- intersect(colnames(SAB08_txi$counts), SAB08_meta$sampleID)

# reorder/filter metadata
SAB08_meta <- SAB08_meta %>%
  dplyr::slice(match(keep_samples, sampleID)) %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# filter txi AND rename columns to unique IDs in one loop
for (elt in intersect(names(SAB08_txi), c("counts", "abundance", "length"))) {
  SAB08_txi[[elt]] <- SAB08_txi[[elt]][, keep_samples, drop = FALSE]
  colnames(SAB08_txi[[elt]]) <- SAB08_meta$sampleID_unique
}

rownames(SAB08_meta) <- SAB08_meta$sampleID_unique

stopifnot(all(colnames(SAB08_txi$counts) == rownames(SAB08_meta)))
```


## merge txi and metasheets
```{r}
#check genes are order
all(rownames(SAB01_txi$counts) == rownames(SAB08_txi$counts),
  rownames(SAB01_txi$counts) == rownames(SAB06_txi$counts))


# combine txi objects
txi_combined <- list(
  counts = cbind(SAB01_txi$counts, SAB08_txi$counts, SAB06_txi$counts),
  abundance = cbind(SAB01_txi$abundance, SAB08_txi$abundance, SAB06_txi$abundance),
  length = cbind(SAB01_txi$length, SAB08_txi$length, SAB06_txi$length),
  countsFromAbundance = SAB01_txi$countsFromAbundance
)

# combine metadata
metadata_combined_clean <- dplyr::bind_rows(
  SAB01_meta,
  SAB08_meta,
  SAB06_meta
)
 
# reorder metadata to match txi columns
metadata_combined_clean <- metadata_combined_clean[
  colnames(txi_combined$counts),
  ,
  drop = FALSE
]

# final safety checks
stopifnot(all(colnames(txi_combined$counts) == rownames(metadata_combined_clean)))
stopifnot(anyDuplicated(colnames(txi_combined$counts)) == 0)
stopifnot(anyDuplicated(rownames(metadata_combined_clean)) == 0)
```

```{r}
# safety checks 
table(metadata_combined_clean$cell_set)
table(metadata_combined_clean$cell_set, metadata_combined_clean$genotype)
table(metadata_combined_clean$cell_set, metadata_combined_clean$treatment)
table(metadata_combined_clean$cell_set, metadata_combined_clean$timepoint)

ncol(txi_combined$counts)
nrow(metadata_combined_clean)
all(colnames(txi_combined$counts) == rownames(metadata_combined_clean))
anyDuplicated(colnames(txi_combined$counts))
anyDuplicated(rownames(metadata_combined_clean))

```

## save
```{r}
saveRDS(txi_combined, file = paste0(rdata_path, "/", exp_name, "_txi.rds"))
saveRDS(metadata_combined_clean, file = paste0(rdata_path, "/", exp_name, "_metadataFinal.rds"))

save_csv(metadata_combined_clean, file.path(data_path, paste0(exp_name, "_metadataFinal.csv")))

```


## Standarize Colors
```{r standarize colors }

# colors fo figures

treatment_colors <- c("CTRL" = "#999999", "RA" = "#D55E00")
genotype_colors  <- c("D21" = "#56B4E9", "T21" = "#CC79A7")

tp_levels <- levels(metadata_combined_clean$timepoint)

timepoint_colors <- setNames(
  ggsci::pal_npg()(length(tp_levels)),
  tp_levels )

annotation_colors <- list(
  treatment = treatment_colors[intersect(names(treatment_colors), unique(metadata_combined_clean$treatment))],
  genotype  = genotype_colors[intersect(names(genotype_colors),  unique(metadata_combined_clean$genotype))],
  timepoint = timepoint_colors[intersect(names(timepoint_colors), tp_levels)]
)

#save
saveRDS(annotation_colors, file = paste0(rdata_path, "/", exp_name, "_annotation_colors.rds"))
```

# -----------------------
# STEP 1: Library / Sequencing QC

## library size and sequencing depth 
```{r}
libqc_dir <- file.path(save_dir, "Library QC")
make_subdir(libqc_dir)

metadata <- metadata_combined_clean
txi <- txi_combined

# load counts and align to metadata 
count_mat <- txi$counts
count_mat <- count_mat[, rownames(metadata)]

# Library size per sample
lib_size <- colSums(count_mat)
qc_counts <- data.frame(
  sampleID_unique = names(lib_size),
  library_size = as.numeric(lib_size),
  detected_genes_ge10 = colSums(count_mat >= 10),
  detected_genes_ge1  = colSums(count_mat >= 1)
) %>%
left_join(
  metadata,
  by = "sampleID_unique"
)

# summary 
print(qc_counts %>%
        select(sampleID_unique, genotype, treatment, timepoint, experimentalBatch,
               library_size, detected_genes_ge10))

base_theme <- theme_bw() +
  theme(
    axis.text.y  = element_text(size = 18, face = "bold"),  
    axis.text.x  = element_text(size = 20),
    axis.title   = element_text(size = 20, face = "bold"),
    plot.title   = element_text(size = 20, face = "bold")
  )

# Plot library size
p1 <- ggplot(qc_counts, aes(x = sampleID_unique, y = library_size)) +
  geom_col() +
  coord_flip() +
  labs(title = "Library size (sum counts)", x = NULL, y = "Total counts")+
  base_theme
ggsave(file.path(libqc_dir, paste0(exp_name, "_QC_library_size.png")), p1, width = 10, height = 6)

# Plot detected genes
p2 <- ggplot(qc_counts, aes(x = sampleID_unique, y = detected_genes_ge10)) +
  geom_col() +
  coord_flip() +
  labs(title = "Detected genes (counts >= 10)", x = NULL, y = "# genes")+
  base_theme
ggsave(file.path(libqc_dir, paste0(exp_name, "_QC_detected_genes_ge10.png")), p2, width = 10, height = 6)

# Optional: log-scale library size (easier to spot low-depth outliers)
p3 <- ggplot(qc_counts, aes(x = sampleID_unique, y = library_size)) +
  geom_col() +
  coord_flip() +
  scale_y_log10() +
  labs(title = "Library size (log10)", x = NULL, y = "Total counts (log10)")+
  base_theme
ggsave(file.path(libqc_dir, paste0(exp_name, "_QC_library_size_log10.png")), p3, width = 10, height = 6)

#counts
qc_counts %>%
  group_by(cell_set) %>%
  summarise(
    mean_lib = mean(library_size),
    median_lib = median(library_size),
    mean_genes = mean(detected_genes_ge10)
  )
```

```{r}
ggplot(
  qc_counts,
  aes(cell_set, library_size)
) +
  geom_boxplot()


ggplot(
  qc_counts,
  aes(cell_set, detected_genes_ge10)
) +
  geom_boxplot()
```


# --------------------
# STEP 2: Check variation

## dds ~1 design
```{r}
dds <- DESeqDataSetFromTximport(
  txi_combined,
  colData = metadata_combined_clean,
  design = ~ 1
)

vsd <- vst(dds, blind = TRUE)

```

## PCA
```{r}
all_sampleIDs <- colnames(assay(vsd))

# PCA 1
save_prefix <- file.path(libqc_dir, paste0(exp_name, "_QC_time_cell_set"))
generatePCA(vsd = assay(vsd), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "cell_set", label = "sampleID_unique",
            save_prefix = save_prefix)
# PCA 2
save_prefix <- file.path(libqc_dir, paste0(exp_name, "_QC_cell_set"))
generatePCA(vsd = assay(vsd), metadata = metadata, all_sampleIDs,
            colby = "cell_set", shape = NULL, label = "timepoint",
            save_prefix = save_prefix)

# PCA 3
save_prefix <- file.path(libqc_dir, paste0(exp_name, "_QC_time_geno"))
generatePCA(vsd = assay(vsd), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "genotype", label = NULL,
            save_prefix = save_prefix)
```


# --------------------
# STEP 2: Build DESeq2 object + run DESeq
```{r}
metadata$cell_set  <- factor(metadata$cell_set, levels = c("Set3", "Set4"))
metadata$genotype  <- factor(metadata$genotype, levels = c("D21", "T21"))
metadata$treatment <- factor(metadata$treatment, levels = c("CTRL", "RA"))
metadata$timepoint <- factor(metadata$timepoint, levels = c("D3", "D4", "D6", "D8", "D10"))

metadata <- metadata[colnames(txi$counts), , drop = FALSE]
stopifnot(all(colnames(txi$counts) == rownames(metadata)))
```

##  DESeq2 global effect 
```{r }
dds <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ cell_set + timepoint + genotype + treatment
)

keep <- edgeR::filterByExpr(
  counts(dds),
  group = interaction(
    metadata$cell_set,
    metadata$timepoint,
    metadata$genotype,
    metadata$treatment,
    drop = TRUE
  )
)

cat("\n============================\n")
cat(" filterByExpr summary\n")
cat("============================\n")
cat("Total genes:    ", nrow(dds), "\n")
cat("Genes kept:     ", sum(keep), "\n")
cat("Genes removed:  ", nrow(dds) - sum(keep), "\n")
cat("Percent removed:", round(100 * (nrow(dds) - sum(keep)) / nrow(dds), 2), "%\n")

dds <- dds[keep, ]

processed_dds <- DESeq(dds)

resultsNames(processed_dds)

```


# --------------------
# STEP 3: PCA - global data check
## Compute VST for PCA/heatmaps

```{r VST}
pca_path <- file.path(save_dir, "PCA")
make_subdir(pca_path)
```

## QC Global PCA  
```{r PCA check}
vsd_dds <- vst(processed_dds, blind = FALSE)

all_sampleIDs <- colnames(assay(vsd_dds))

# PCA 1: biology-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_colTime"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "cell_set", label = "genotype",
            save_prefix = save_prefix )

# PCA 2: QC-set-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_cell_set"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "cell_set", shape = "genotype", label = NA,
            save_prefix = save_prefix, scale_to_axes = TRUE)

# PCA 3: 
save_prefix <- file.path(pca_path, paste0(exp_name, "_time_geno"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "genotype", label = NA,
            save_prefix = save_prefix, scale_to_axes = TRUE)

```


```{r PCA check}
pcaLoadings(
  vsd_data = assay(vsd_dds),metadata = metadata, sample_ids = all_sampleIDs,
  exp_name = exp_name, name = "allSamples", results_dir = pca_path)

# print variance captured by each PCA 
p <- PCAtools::pca(assay(vsd_dds), metadata = metadata, removeVar = 0.9)
round(p$variance, 2)[1:6]
```


###  - OPTIONAL: Limma Batch Correction

```{r limma}
#preserve design so doesnt correct for biology 
design_mat <- model.matrix(~ genotype_timepoint_treatment, data = colData(vsd_dds))

vsd_dds_limmaCorrected <- limma::removeBatchEffect(
  assay(vsd_dds),
  batch = vsd_dds$experimentalBatch,
  design = design_mat
)

# Ensure limma-corrected matrix has sample IDs and is ordered like metadata
colnames(vsd_dds_limmaCorrected) <- colnames(assay(vsd_dds))
vsd_dds_limmaCorrected <- vsd_dds_limmaCorrected[, rownames(metadata), drop = FALSE]

saveRDS(vsd_dds_limmaCorrected, file = paste0(rdata_path, "/", exp_name, "_vsd_dds_limmaCorrected.rds"))

# recheck PCA
all_sampleIDs <- colnames(vsd_dds_limmaCorrected)

# PCA 1: biology-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_limma_allSamples_colTime"))
generatePCA(vsd = vsd_dds_limmaCorrected, metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "treatment", label = NA,
            save_prefix = save_prefix)

# PCA 2: QC-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_limma_allSamples_colBatch"))
generatePCA(vsd = vsd_dds_limmaCorrected, metadata = metadata, all_sampleIDs,
            colby = "experimentalBatch", shape = "genotype", label = NA,
            save_prefix = save_prefix)

pcaLoadings(
  vsd_data = vsd_dds_limmaCorrected, metadata = metadata, sample_ids = all_sampleIDs,
  exp_name = exp_name, name = "allSamples_limmaCorrection", results_dir = pca_path)  


# print variance captured by each PCA 
p_lima <- PCAtools::pca(vsd_dds_limmaCorrected, metadata = metadata, removeVar = 0.9)
round(p_lima$variance, 2)[1:6]
```

limmas corrected not necessary but doesnt appear to mess with biology so will continue to use for cleaner PCAs 

### PC2, 3, 4
```{r}

# PC1 vs PC3
p13_time <- PCAtools::biplot(p, x = "PC1", y = "PC3", 
                             colby = "cell_set", shape = "treatment", lab = metadata[["timepoint"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PCA_PC1_PC3_cell_set.png")),
       p13_time, width = 6, height = 5, dpi = 300)

# PC2 vs PC3
p23_time <- PCAtools::biplot(p, x = "PC2", y = "PC3",
                            colby = "cell_set", shape = "treatment", lab = metadata[["timepoint"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC2_PC3_cell_set.png")),
       p23_time, width = 6, height = 5, dpi = 300)


p23_time <- PCAtools::biplot(p, x = "PC2", y = "PC3",
                            colby = "timepoint", shape = "treatment", lab = metadata[["genotype"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC2_PC3_timepoint.png")),
       p23_time, width = 6, height = 5, dpi = 300)

# PC1 vs PC4
p14_batch <- PCAtools::biplot(p, x = "PC1", y = "PC4",
                             colby = "cell_set", shape = "treatment", lab = metadata[["timepoint"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC1_PC4_cell_set.png")),
       p14_batch, width = 6, height = 5, dpi = 300)

p14_batch <- PCAtools::biplot(p, x = "PC1", y = "PC4",
                             colby = "timepoint", shape = "cell_set", lab = metadata[["treatment"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC1_PC4_tim.png")),
       p14_batch, width = 6, height = 5, dpi = 300)


```

### QC PCAs
```{r}
# D3 only
d3_ids <- rownames(
  metadata %>%
    dplyr::filter(timepoint == "D3")
)

vsd_d3 <- assay(vsd_dds)[, d3_ids]
meta_d3 <- metadata[d3_ids, , drop = FALSE]

generatePCA(vsd_d3, meta_d3, d3_ids,
            colby="cell_set", shape="genotype", label=NULL,
            save_prefix=file.path(pca_path, paste0(exp_name, "_QC_D3_only")),
            annotation_colors=annotation_colors)

# D3, D4, D6
early_ids <- rownames(
  metadata %>%
    dplyr::filter(timepoint %in% c("D3", "D4", "D6"))
)

vsd_early <- assay(vsd_dds)[, early_ids]
meta_early <- metadata[early_ids, , drop = FALSE]

generatePCA(vsd_early, meta_early, early_ids,
  colby = "timepoint", shape = "cell_set", label = "genotype",
  save_prefix = file.path(pca_path, paste0(exp_name, "_QC_D3_D4_D6_timepoint")),
  annotation_colors = annotation_colors)

generatePCA(vsd_early, meta_early, early_ids,
  colby = "treatment", shape = "cell_set", label = "genotype",
  save_prefix = file.path(pca_path, paste0(exp_name, "_QC_D3_D4_D6_treatment")),
  annotation_colors = annotation_colors
)

generatePCA(vsd_early, meta_early, early_ids,
  colby = "genotype", shape = "cell_set", label = "treatment",
  save_prefix = file.path(pca_path, paste0(exp_name, "_QC_D3_D4_D6_genotype")),
  annotation_colors = annotation_colors
)

```

# --  STOPPED HERE -- #

## PCAs
```{r}

# all samples
ids <- get_ids(metadata)
generatePCA(vsd_mat, metadata, ids,
            colby="timepoint", shape="treatment", label="genotype",
            save_prefix=file.path(pca_path, paste0(exp_name, "_allSamples")),
            annotation_colors=annotation_colors)

generatePCA(vsd_mat, metadata, ids,
            colby="timepoint", shape="genotype", label="treatment",
            save_prefix=file.path(pca_path, paste0(exp_name, "_allSamples2")),
            annotation_colors=annotation_colors)


# CTRL only
ids <- get_ids(metadata, treatments="CTRL")

generatePCA(vsd_mat, metadata, ids,
            colby="timepoint", shape="genotype", label=NA,
            save_prefix=file.path(pca_path, paste0(exp_name, "_CTRL_only")),
            annotation_colors=annotation_colors)


# RA only + D3 pretreatment
ids <- metadata %>%
  dplyr::filter(treatment == "RA" | timepoint == "D3") %>%
  dplyr::pull(sampleID)

generatePCA(vsd_mat, metadata, ids,
            colby="timepoint", shape="genotype", label=NA,
            save_prefix=file.path(pca_path, paste0(exp_name, "_RA_only")),
            annotation_colors=annotation_colors)

```

## Correlation heatmaps 
(using VST or batch-corrected)

### all samples
```{r}

heatmap_path <- file.path(save_dir, "Heatmap")
make_subdir(heatmap_path)


heatmap_plots <- list()

# SORT:timepoint -> treatment -> genotype
genotype_order <- factor(metadata$genotype, levels = c("D21", "T21"))
meta_hm <- metadata[order(metadata$timepoint, metadata$treatment, genotype_order), ]

# subset VST matrix using defined order
vsd_all <- vsd_mat[, meta_hm$sampleID, drop = FALSE]
cor_matrix <- cor(vsd_all)

rownames(cor_matrix) <- meta_hm$genotype_timepoint_treatment_batch
colnames(cor_matrix) <- meta_hm$sampleID


col_annotation <- data.frame(
  timepoint = meta_hm$timepoint,
  treatment = meta_hm$treatment,
  genotype  = meta_hm$genotype
)
rownames(col_annotation) <- meta_hm$sampleID

# gap based on num of timepoints 
num_clusters <- length(unique(meta_hm$timepoint))

# generate heatmap 
heatmap_filename <- file.path(heatmap_path, paste0(exp_name, "_heatmap_allSamples.png"))
ph1 <- pheatmap(
  cor_matrix,
  filename = heatmap_filename,
  main = "Correlation Heatmap - All Samples",
  annotation_col = col_annotation,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  cutree_rows = num_clusters,
  display_numbers = FALSE,
  color = viridis::viridis(100),
  annotation_colors = annotation_colors,
  border_color = NA,
  fontsize_row = 10,
  fontsize_col = 8,
  height = 10,
  width = 10
)

heatmap_plots[["heatmap_allSamples"]] <- ph1
```

### by treatment 
```{r}
unique_treatments <- unique(metadata$treatment)

# loop through each treatent and generate heatpmap
for (trt in unique_treatments) {

  #sort for plotting
  meta_sub <- metadata[metadata$treatment == trt, ]
  genotype_order <- factor(meta_sub$genotype, levels = c("D21", "T21"))
  meta_sub <- meta_sub[order(meta_sub$timepoint, genotype_order), ]

  vsd_subset <- vsd_mat[, meta_sub$sampleID, drop = FALSE]
  cor_matrix <- cor(vsd_subset)

  rownames(cor_matrix) <- meta_sub$genotype_timepoint_treatment_batch
  colnames(cor_matrix) <- meta_sub$sampleID

  col_annotation <- data.frame(
    timepoint = meta_sub$timepoint,
    genotype  = meta_sub$genotype
  )
  rownames(col_annotation) <- meta_sub$sampleID

  num_clusters <- length(unique(meta_sub$timepoint))
  heatmap_filename <- file.path(heatmap_path, paste0(exp_name, "_heatmap_", trt, ".png"))

  phtrt <- pheatmap(
    cor_matrix,
    filename = heatmap_filename,
    main = paste("Correlation Heatmap -", trt, "Samples"),
    annotation_col = col_annotation,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    cutree_rows = num_clusters,
    display_numbers = FALSE,
    color = viridis::viridis(100),
    annotation_colors = annotation_colors,
    border_color = NA,
    fontsize_row = 10,
    fontsize_col = 8,
    height = 10,
    width = 10
  )

  heatmap_plots[[paste0("heatmap_", trt)]] <- phtrt
}

# save all 
saveRDS(heatmap_plots, file = file.path(rdata_path, paste0(exp_name, "_all_heatmap_objects.rds")))

```


## DESeq RA interaction
```{r}
dds_int <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ cell_set + timepoint + genotype + treatment + genotype:treatment
)

dds_int <- DESeq(dds_int)
resultsNames(dds_int)
```


## --------------------
# STEP 4: DEseq2 DEG analysis

```{r}
log2FC_Cutoff <- 0.5
fdrCutoff <- 0.05

deg_path <- file.path(save_dir, paste0("L2FC", log2FC_Cutoff, "_FDR", fdrCutoff))
make_subdir(deg_path)

# Check metadata and DESeq object alignment
if (!all(rownames(metadata) == colnames(processed_dds))) {
  warning("⚠️ Rownames in metadata do NOT match colnames in processed_dds!")
  
  cat("🔍 Mismatch summary:\n")
  cat(" - Metadata rows:", nrow(metadata), "\n")
  cat(" - DESeq2 columns:", ncol(processed_dds), "\n")
  cat(" - First few metadata row names:\n", head(rownames(metadata)), "\n")
  cat(" - First few DESeq2 column names:\n", head(colnames(processed_dds)), "\n")
  
} else {
  cat("✅ Metadata rownames match DESeq2 column names.\n")
}

```


##  DE Analysis
-perform differential expression analysis for comparision across time points
-apply the DESeq2 package to extract differentially expressed genes (DEGs) and organizes in two list (a) filtered DEGS, up dn (b) unfilterd all genes 

```{r DEG comps}
# define DEG comparisons to make 
deg_comparisons <- list(
  "T21_CTRL_vs_D21_CTRL"   = list(genotype1 = "T21", genotype2 = "D21", treat1 = "CTRL", treat2 = "CTRL"),
  "T21_RA_vs_D21_RA"   = list(genotype1 = "T21", genotype2 = "D21", treat1 = "RA", treat2 = "RA"),
  "D21_RA_vs_D21_CTRL" = list(genotype = "D21", treat1 = "CTRL", treat2 = "RA"),
  "T21_RA_vs_T21_CTRL" = list(genotype = "T21", treat1 = "CTRL", treat2 = "RA")
)

timepoints <- unique(metadata$timepoint)

```


```{r DE analysis, message=FALSE}
# master lists 
filtered_results <- list()
unfiltered_results <- list()

for (comp in names(deg_comparisons)) {
  comp_info <- deg_comparisons[[comp]]
  
  # lists for current comparison
  filtered_list <- list()
  unfiltered_list <- list()
  
  for (d in timepoints) {
    
    # Dynamically build contrast groups and ID
    if (!is.null(comp_info$genotype1)) {
      x <- paste0(comp_info$genotype1, "_", d, "_", comp_info$treat1)
      y <- paste0(comp_info$genotype2, "_", d, "_", comp_info$treat2)
      id <- paste0(comp_info$genotype1, "_", d, "_", comp_info$treat1, 
                   "_vs_", comp_info$genotype2, "_", d, "_", comp_info$treat2)
    } else {
      genotype <- comp_info$genotype
      x <- paste0(genotype, "_", d, "_", comp_info$treat2)
      y <- paste0(genotype, "_", d, "_", comp_info$treat1)
      id <- paste0(genotype, "_", d, "_", comp_info$treat2, 
                   "_vs_", genotype, "_", d, "_", comp_info$treat1)
    }
    
    # Only run if both contrast groups exist
    if (x %in% unique(processed_dds$genotype_timepoint_treatment) && 
        y %in% unique(processed_dds$genotype_timepoint_treatment)) {
      
      res <- DESeq2::results(processed_dds, contrast = c("genotype_timepoint_treatment", x, y), alpha=fdrCutoff)
      res_shrunk <- lfcShrink(dds = processed_dds, contrast = c("genotype_timepoint_treatment", x, y), res = res, type = "ashr")
      
      # Add both raw and shrunk LFC using gene_id join
      res_df <- as.data.frame(res) %>% rownames_to_column("gene_id")
      res_shrunk_df <- as.data.frame(res_shrunk) %>% rownames_to_column("gene_id")
      
      # Join by gene_id and filter out rows with NA padj
      merged_df <- inner_join(res_shrunk_df, res_df[, c("gene_id", "log2FoldChange")], by = "gene_id") %>%
        filter(!is.na(padj)) %>%
        dplyr::rename(
          log2FoldChange_shrunk = log2FoldChange.x,
          log2FoldChange_raw = log2FoldChange.y
        ) %>%
        dplyr::select(gene_id, baseMean, log2FoldChange_shrunk, log2FoldChange_raw, pvalue, padj)
      
      unfiltered_list[[id]] <- merged_df

      # Filter significant DEGs (based on raw LFC)
      tmp <- merged_df %>%
        dplyr::filter(!is.na(padj), abs(log2FoldChange_raw) > log2FC_Cutoff, padj < fdrCutoff)

      filtered_list[[paste0(id, ".up")]] <- tmp %>% filter(log2FoldChange_raw > 0) %>% pull(gene_id)
      filtered_list[[paste0(id, ".dn")]] <- tmp %>% filter(log2FoldChange_raw < 0) %>% pull(gene_id)
    }
  }
  
  # Save per-comparison lists 
  saveRDS(filtered_list, file = file.path(rdata_path, paste0(exp_name, "_DEG_filtered_", comp, "_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".rds")))
  saveRDS(unfiltered_list, file = file.path(rdata_path, paste0(exp_name, "_DEG_unfiltered_", comp, ".rds")))
  
  # Export Excel of unfiltered data
  results_for_excel <- lapply(unfiltered_list, function(df) data.frame(Gene = df$gene_id, df[ , -1]))
  write_xlsx(results_for_excel, path = file.path(deg_path, paste0(exp_name, "_DEG_unfiltered_", comp, ".xlsx")))
  
  # Store into master result containers
  filtered_results[[comp]] <- filtered_list
  unfiltered_results[[comp]] <- unfiltered_list
}


saveRDS(filtered_results, file = file.path(rdata_path, paste0(exp_name, "_DEG_filtered_ALL_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".rds")))
saveRDS(unfiltered_results, file = file.path(rdata_path, paste0(exp_name, "_DEG_unfiltered_ALL.rds")))
```


## DEG Summary 
```{r}
deg_summary <- list()

for (comparison in names(filtered_results)) {
  contrast_list <- filtered_results[[comparison]]
  if (length(contrast_list) == 0) next
  
  for (contrast_id in names(contrast_list)) {
    direction <- ifelse(grepl("\\.up$", contrast_id), "Upregulated", "Downregulated")
    contrast_base <- sub("\\.(up|dn)$", "", contrast_id)
    deg_summary[[length(deg_summary) + 1]] <- data.frame(
      Comparison = comparison,
      Contrast   = contrast_base,
      Direction  = direction,
      NumGenes   = length(contrast_list[[contrast_id]])
    )
  }
}

deg_summary_df <- bind_rows(deg_summary) %>%
  arrange(Comparison, Contrast, Direction)

# Total DEG count across all contrasts
total_degs <- sum(deg_summary_df$NumGenes, na.rm = TRUE)

# CSV
save_csv(deg_summary_df, path = file.path(deg_path, paste0(exp_name, "_DEG_summary_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".csv")))

# PDF 
pdf_path <- file.path(deg_path, paste0(exp_name, "_DEG_summary_log2FC.", log2FC_Cutoff,
         "_FDR.", fdrCutoff, ".pdf"))

title_text <- paste0("DEG Summary | log2FC cutoff = ", log2FC_Cutoff,
                     " | FDR cutoff = ", fdrCutoff,
                     " | Total DEGs = ", total_degs)

pdf(pdf_path, width = 8.5, height = 11)
grid.text(title_text, gp = gpar(fontsize = 16, fontface = "bold"), x = 0.5, y = 0.97)  # centered top
gridExtra::grid.table(deg_summary_df)

dev.off()
```




## Metascape Export 
```{r}
meta_path <- file.path(deg_path, "Metascape")
make_subdir(meta_path)

#global background
all_tested_global <- unique(unlist(
  lapply(unfiltered_results, function(comp_list) {
    lapply(comp_list, function(df) df$gene_id)
  })
))

for (comp in names(filtered_results)) {
  comp_lists <- filtered_results[[comp]]   # named vectors for each contrast .up/.dn
  if (length(comp_lists) == 0) next

  # sanitize: ensure character, drop NA/dup
  comp_lists <- lapply(comp_lists, function(v) unique(na.omit(as.character(v))))

  # Global background (default):
  comp_lists[["_BACKGROUND"]] <- all_tested_global

  # Rectangle-ize to a data frame (pad shorter columns with NA)
  max_len <- max(lengths(comp_lists))
  df_comp <- tibble::tibble(.rows = max_len)
  for (nm in names(comp_lists)) {
    vec <- comp_lists[[nm]]
    length(vec) <- max_len
    df_comp[[nm]] <- vec
  }

  out_csv <- file.path(meta_path, paste0(exp_name, "_", comp, "_for_Metascape.csv"))
  readr::write_csv(df_comp, out_csv)
}
```


## --------------------
# STEP 5: Chromosome Annotation
Use same bed file that was used to map reads to annotate chromosomes
- bed file give you chromosome -> tx_id 
- txi gives you tx_id ->gene_id 


## Chromsome Map 
### Read bed file - chromosome to tx_id mapping then tx_id to gene_id 
```{r}
chrom_dir <- file.path(save_dir, paste0("Chromosome Annotations"))
make_subdir(chrom_dir)

bed_cols <- c("chrom","chromStart","chromEnd","tx_id",
              "score","strand","thickStart","thickEnd",
              "itemRgb","blockCount","blockSizes","blockStarts")

genes_bed_raw <- readr::read_tsv(
  file.path(data_path, "genes.bed"),
  col_names = bed_cols,
  comment   = "#",
  show_col_types = FALSE
)

# normalize chromosome labels 
standard_chrs <- c(paste0("chr", 1:22), "chrX", "chrY")
norm_chr <- function(x) {
  x <- as.character(x)
  x <- ifelse(grepl("^chr", x), x, paste0("chr", x)) # add 'chr' if missing
  x
}

genes_bed <- genes_bed_raw %>%
  mutate(chrom = norm_chr(chrom)) %>%
  filter(chrom %in% standard_chrs) %>%
  distinct(tx_id, chrom)

# read tsv file - tx_id to gene_id mapping
tx2gene <- readr::read_tsv(
  file.path(star_salmon_path, "tx2gene.tsv"),
  col_names = c("tx_id","gene_id"),
  show_col_types = FALSE
) %>%
  distinct(tx_id, gene_id)


#check both 
cat("tx2gene rows:", nrow(tx2gene), 
    " unique tx:", dplyr::n_distinct(tx2gene$tx_id), "\n")
cat("tx in BED & tx2gene:", sum(tx2gene$tx_id %in% genes_bed$tx_id), "\n")
```

### Build final gene to chrom map
```{r}
gene_chrom <- genes_bed %>%
  dplyr::select(tx_id, chrom) %>%
  dplyr::left_join(tx2gene, by = "tx_id") %>%
  dplyr::select(gene_id, chrom) %>%
  dplyr::distinct()

saveRDS(gene_chrom,
  file = file.path(rdata_path, paste0(exp_name, "_gene_to_chrom.rds")))

#-- Check genes map to chromosome map --
# load DEGS
unfiltered_results <- readRDS( file.path(rdata_path, paste0(exp_name, "_DEG_unfiltered_ALL.rds")))
filtered_results <- readRDS( file.path(rdata_path, paste0(exp_name, "_DEG_filtered_ALL_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".rds")))

# full tested gene universe (unique, non-NA, character)
all_tested <- unique(na.omit(as.character(unlist(
  lapply(unfiltered_results, function(comp_list) {
    lapply(comp_list, function(df) df$gene_id)
  })
))))

# check that all genes mapped to chromosomes
all_degs <- unique(na.omit(as.character(unlist(filtered_results))))
mapped_tested <- length(intersect(all_tested, gene_chrom$gene_id))
mapped_degs   <- length(intersect(all_degs,   gene_chrom$gene_id))

cat(sprintf("Tested genes: %d | Mapped: %d (%.1f%%)\n",
          length(all_tested), mapped_tested, 100 * mapped_tested / max(1, length(all_tested))))
cat(sprintf("DEGs total:   %d | Mapped: %d (%.1f%%)\n",
          length(all_degs), mapped_degs, 100 * mapped_degs / max(1, length(all_degs))))
```

## Annotate DEG list 
```{r }
# Annotate DEG list 
chromosome_results <- lapply(filtered_results, function(day_list) {
  lapply(day_list, function(gene_vec) {
    tibble(gene_id = as.character(gene_vec)) %>%
      dplyr::left_join(gene_chrom, by = "gene_id")
  })
})

```


### Plot DEGS/ Chrom
```{r}
chrom_plots <- list()

# 1. summarize
chr_counts_df <- summarize_chr_counts(chromosome_results, 
                                      gene_chrom = gene_chrom, all_tested = all_tested )

# 2. clean up titles
chr_counts_df_pretty <- prettify_chr_counts(chr_counts_df)

# 3. plot with shared axis 
chrom_plots$facet_updown <- plot_chr_updown( chr_counts_df_pretty,
  save_dir = chrom_dir, mode = "facet")

# 4. up/down indiv graphs by day
chrom_plots$updown_day <- plot_chr_updown( chr_counts_df_pretty, save_dir = chrom_dir, mode = "individual")
```


## Total DEGs per chrosmome/day
```{r}

total_ratio_df <- summarize_total_ratio(chr_counts_df)
chrom_plots$total_byday <-plot_total_ratio_by_chrom_and_day(total_ratio_df, save_dir = chrom_dir, annotation_colors = annotation_colors)

```

## Save all
```{r}

saveRDS( chrom_plots,
  file = file.path(rdata_path, paste0(exp_name, "_chromosome_plots.rds")))

saveRDS(chr_counts_df_pretty, file = file.path(rdata_path, paste0(exp_name, "_chr_counts_pretty.rds")))

saveRDS(total_ratio_df, file = file.path(rdata_path, paste0(exp_name, "_total_ratio_df.rds")))

```


## Check for sex 
```{r}
sex_qc <- make_sex_qc(vsd_mat, metadata, line_name = exp_name)

sex_qc$plot
sex_qc$predicted_sex

outfile <- file.path(chrom_dir, paste0(exp_name, "_sexQC.pdf"))

ggsave(outfile, sex_qc$plot, width = 6, height = 5, dpi = 300)

```

