---
title: "SET3 - SET4 Comparison"
params:                              
  experiment_id: S3_S4_COMP
  run_date: 2026-06-08
  author: "Shalise Burch"
output: html_document  
knitr:
  opts_chunk:
    echo: TRUE        
    warning: FALSE     
    message: FALSE     
---

STEP 0: Set up


# ------------------
# STEP 0: Setup
```{r libraries and variables, message=FALSE, warning=FALSE}
#UPDATED:
# 2026-06-08 14:30 SAB ------------------------------

library(here)
library(DESeq2)
library(limma)
library(PCAtools)
library(tximport)
library(pheatmap)
library(dplyr) 
library(ggplot2)
library(gtools)
library(ggsci)
library(stringr) 
library(edgeR)
library(grid)
library(data.table)
library(tidyverse)
library(writexl)
library(ashr)


exp_name <- params$experiment_id
data_path <- here("rnaseq", "01_data")
results_path <- here("rnaseq", "02_results")
rdata_path <- here("rnaseq", "03_rdata")
source(here("rnaseq", "00_scripts", "rnaseq_functions.R"))

colors <- colorRampPalette(c("#3B4CC0", "white", "#B40426"))(100)
breaks <- seq(-2, 2, length.out = 101)

# load metasheets and txi
# -- SET 3 -- 
SAB01_meta <- readRDS("/Users/shalise/Library/CloudStorage/OneDrive-TheUniversityofChicago/R_Studio/SAB01/rnaseq/03_rdata/SAB01_metadataFinal.rds")

SAB01_star_salmon_path <- "/Volumes/project/imoskowitz/shared/sequencing.processed/20240412_LH00315_0161_B222TJLLT1-XRS-Pl21-IM-RS54-hg38/star_salmon"


# -- SET 4 -- 
SAB08_meta <- readRDS("/Users/shalise/Library/CloudStorage/OneDrive-TheUniversityofChicago/R_Studio/SAB08/rnaseq/03_rdata/SAB08_metadataFinal_outliers_removed.rds")

SAB08_star_salmon_path <- "/Volumes/project/imoskowitz/shared/sequencing.processed/SAB08/star_salmon"


save_dir <- file.path(results_path, "S3_S4_COMP")
make_subdir(save_dir)

```


## Clean metasheet
```{r clean data}
# add seq_run to SAB01 to match SAB08
SAB01_meta$seq_run <- "1"

# add cell set identifier
SAB01_meta$cell_set <- "Set3"
SAB08_meta$cell_set <- "Set4"


# add new sampleID identifier
SAB01_meta <- SAB01_meta %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

SAB08_meta <- SAB08_meta %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# Make sure rownames are new sample IDs
rownames(SAB01_meta) <- SAB01_meta$sampleID_unique
rownames(SAB08_meta) <- SAB08_meta$sampleID_unique


#check they match 
setdiff(colnames(SAB01_meta), colnames(SAB08_meta))
setdiff(colnames(SAB08_meta), colnames(SAB01_meta))


```



## Load txi object
```{r txi}

# load txi
SAB01_txi <- salmon_import(SAB01_star_salmon_path)
SAB08_txi <- salmon_import(SAB08_star_salmon_path)

#check sample names 
colnames(SAB01_txi$counts)
rownames(SAB01_meta)

colnames(SAB08_txi$counts)
rownames(SAB08_meta)
```


## clean txi names
### set 3
```{r clean txi names}

# Set3: strip SAB01 prefix, align metadata, then assign unique IDs
strip_prefix <- function(x) sub("^SAB01_", "", x)

# strip first
for (elt in intersect(names(SAB01_txi), c("counts", "abundance", "length"))) {
  colnames(SAB01_txi[[elt]]) <- strip_prefix(colnames(SAB01_txi[[elt]]))
}

# now align metadata to stripped txi column order
SAB01_meta <- SAB01_meta %>%
  dplyr::slice(match(colnames(SAB01_txi$counts), sampleID)) %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# rename all txi elements to unique IDs
for (elt in intersect(names(SAB01_txi), c("counts", "abundance", "length"))) {
  colnames(SAB01_txi[[elt]]) <- SAB01_meta$sampleID_unique
}

rownames(SAB01_meta) <- SAB01_meta$sampleID_unique

stopifnot(all(colnames(SAB01_txi$counts) == rownames(SAB01_meta)))
```

### set4
```{r}

# strip SAB08 prefix
strip_prefix <- function(x) sub("^SAB08_", "", x)

for (elt in intersect(names(SAB08_txi), c("counts", "abundance", "length"))) {
  colnames(SAB08_txi[[elt]]) <- strip_prefix(colnames(SAB08_txi[[elt]]))
}

# samples shared between txi and metadata
keep_samples <- intersect(colnames(SAB08_txi$counts), SAB08_meta$sampleID)

# reorder/filter metadata
SAB08_meta <- SAB08_meta %>%
  dplyr::slice(match(keep_samples, sampleID)) %>%
  dplyr::mutate(
    sampleID_original = sampleID,
    sampleID_unique = paste(cell_set, sampleID, sep = "_")
  )

# filter txi AND rename columns to unique IDs in one loop
for (elt in intersect(names(SAB08_txi), c("counts", "abundance", "length"))) {
  SAB08_txi[[elt]] <- SAB08_txi[[elt]][, keep_samples, drop = FALSE]
  colnames(SAB08_txi[[elt]]) <- SAB08_meta$sampleID_unique
}

rownames(SAB08_meta) <- SAB08_meta$sampleID_unique

stopifnot(all(colnames(SAB08_txi$counts) == rownames(SAB08_meta)))
```


## merge txi and metasheets
```{r}
# combine txi objects
txi_combined <- list(
  counts = cbind(SAB01_txi$counts, SAB08_txi$counts),
  abundance = cbind(SAB01_txi$abundance, SAB08_txi$abundance),
  length = cbind(SAB01_txi$length, SAB08_txi$length),
  countsFromAbundance = SAB01_txi$countsFromAbundance
)

# combine metadata
metadata_combined_clean <- dplyr::bind_rows(
  SAB01_meta,
  SAB08_meta
)

# reorder metadata to match txi columns
metadata_combined_clean <- metadata_combined_clean[
  colnames(txi_combined$counts),
  ,
  drop = FALSE
]

# final safety checks
stopifnot(all(colnames(txi_combined$counts) == rownames(metadata_combined_clean)))
stopifnot(anyDuplicated(colnames(txi_combined$counts)) == 0)
stopifnot(anyDuplicated(rownames(metadata_combined_clean)) == 0)
```

```{r}
# safety checks 
table(metadata_combined_clean$cell_set)
table(metadata_combined_clean$cell_set, metadata_combined_clean$genotype)
table(metadata_combined_clean$cell_set, metadata_combined_clean$treatment)
table(metadata_combined_clean$cell_set, metadata_combined_clean$timepoint)

ncol(txi_combined$counts)
nrow(metadata_combined_clean)
all(colnames(txi_combined$counts) == rownames(metadata_combined_clean))
anyDuplicated(colnames(txi_combined$counts))
anyDuplicated(rownames(metadata_combined_clean))

```

## save
```{r}
saveRDS(txi_combined, file = paste0(rdata_path, "/", exp_name, "_txi.rds"))
saveRDS(metadata_combined_clean, file = paste0(rdata_path, "/", exp_name, "_metadataFinal.rds"))

save_csv(metadata_combined_clean, file.path(data_path, paste0(exp_name, "_metadataFinal.csv")))

```


## Standarize Colors
```{r standarize colors }

# colors fo figures

treatment_colors <- c("CTRL" = "#999999", "RA" = "#D55E00")
genotype_colors  <- c("D21" = "#56B4E9", "T21" = "#CC79A7")

tp_levels <- levels(metadata_combined_clean$timepoint)

timepoint_colors <- setNames(
  ggsci::pal_npg()(length(tp_levels)),
  tp_levels )

annotation_colors <- list(
  treatment = treatment_colors[intersect(names(treatment_colors), unique(metadata_combined_clean$treatment))],
  genotype  = genotype_colors[intersect(names(genotype_colors),  unique(metadata_combined_clean$genotype))],
  timepoint = timepoint_colors[intersect(names(timepoint_colors), tp_levels)]
)

#save
saveRDS(annotation_colors, file = paste0(rdata_path, "/", exp_name, "_annotation_colors.rds"))
```

# -----------------------
# STEP 1: Library / Sequencing QC

## library size and sequencing depth 
```{r}
libqc_dir <- file.path(save_dir, "Library QC")
make_subdir(libqc_dir)

metadata <- metadata_combined_clean
txi <- txi_combined

# load counts and align to metadata 
count_mat <- txi$counts
count_mat <- count_mat[, rownames(metadata)]

# Library size per sample
lib_size <- colSums(count_mat)
qc_counts <- data.frame(
  sampleID_unique = names(lib_size),
  library_size = as.numeric(lib_size),
  detected_genes_ge10 = colSums(count_mat >= 10),
  detected_genes_ge1  = colSums(count_mat >= 1)
) %>%
left_join(
  metadata,
  by = "sampleID_unique"
)

# summary 
print(qc_counts %>%
        select(sampleID_unique, genotype, treatment, timepoint, experimentalBatch,
               library_size, detected_genes_ge10))

base_theme <- theme_bw() +
  theme(
    axis.text.y  = element_text(size = 18, face = "bold"),  
    axis.text.x  = element_text(size = 20),
    axis.title   = element_text(size = 20, face = "bold"),
    plot.title   = element_text(size = 20, face = "bold")
  )

# Plot library size
p1 <- ggplot(qc_counts, aes(x = sampleID_unique, y = library_size)) +
  geom_col() +
  coord_flip() +
  labs(title = "Library size (sum counts)", x = NULL, y = "Total counts")+
  base_theme
ggsave(file.path(libqc_dir, paste0(exp_name, "_QC_library_size.png")), p1, width = 10, height = 6)

# Plot detected genes
p2 <- ggplot(qc_counts, aes(x = sampleID_unique, y = detected_genes_ge10)) +
  geom_col() +
  coord_flip() +
  labs(title = "Detected genes (counts >= 10)", x = NULL, y = "# genes")+
  base_theme
ggsave(file.path(libqc_dir, paste0(exp_name, "_QC_detected_genes_ge10.png")), p2, width = 10, height = 6)

# Optional: log-scale library size (easier to spot low-depth outliers)
p3 <- ggplot(qc_counts, aes(x = sampleID_unique, y = library_size)) +
  geom_col() +
  coord_flip() +
  scale_y_log10() +
  labs(title = "Library size (log10)", x = NULL, y = "Total counts (log10)")+
  base_theme
ggsave(file.path(libqc_dir, paste0(exp_name, "_QC_library_size_log10.png")), p3, width = 10, height = 6)

#counts
qc_counts %>%
  group_by(cell_set) %>%
  summarise(
    mean_lib = mean(library_size),
    median_lib = median(library_size),
    mean_genes = mean(detected_genes_ge10)
  )
```

```{r}
ggplot(
  qc_counts,
  aes(cell_set, library_size)
) +
  geom_boxplot()


ggplot(
  qc_counts,
  aes(cell_set, detected_genes_ge10)
) +
  geom_boxplot()
```


# --------------------
# 1. DESeq2 global effects

## ~1 design
```{r}
dds <- DESeqDataSetFromTximport(
  txi_combined,
  colData = metadata_combined_clean,
  design = ~ 1
)

vsd <- vst(dds, blind = TRUE)

```

### PCA
```{r}
all_sampleIDs <- colnames(assay(vsd))

# PCA 1
save_prefix <- file.path(libqc_dir, paste0(exp_name, "_QC_time_cell_set"))
generatePCA(vsd = assay(vsd), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "cell_set", label = "sampleID_unique",
            save_prefix = save_prefix)
# PCA 2
save_prefix <- file.path(libqc_dir, paste0(exp_name, "_QC_cell_set"))
generatePCA(vsd = assay(vsd), metadata = metadata, all_sampleIDs,
            colby = "cell_set", shape = NULL, label = "timepoint",
            save_prefix = save_prefix)

# PCA 3
save_prefix <- file.path(libqc_dir, paste0(exp_name, "_QC_time_geno"))
generatePCA(vsd = assay(vsd), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "genotype", label = NULL,
            save_prefix = save_prefix)
```


# --------------------
# 3. DESeq2 main effects ~ cell_set + timepoint + genotype + treatment
```{r}
main_effects_path <- file.path(save_dir, "Main_Effects")
make_subdir(main_effects_path)


metadata$cell_set  <- factor(metadata$cell_set, levels = c("Set3", "Set4"))
metadata$genotype  <- factor(metadata$genotype, levels = c("D21", "T21"))
metadata$treatment <- factor(metadata$treatment, levels = c("CTRL", "RA"))
metadata$timepoint <- factor(metadata$timepoint, levels = c("D3", "D4", "D6", "D8", "D10"))

metadata <- metadata[colnames(txi$counts), , drop = FALSE]
stopifnot(all(colnames(txi$counts) == rownames(metadata)))
```

##  ~ cell_set + timepoint + genotype + treatment
```{r }
dds <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ cell_set + timepoint + genotype + treatment
)

keep <- edgeR::filterByExpr(
  counts(dds),
  group = interaction(
    metadata$cell_set,
    metadata$timepoint,
    metadata$genotype,
    metadata$treatment,
    drop = TRUE
  )
)

cat("\n============================\n")
cat(" filterByExpr summary\n")
cat("============================\n")
cat("Total genes:    ", nrow(dds), "\n")
cat("Genes kept:     ", sum(keep), "\n")
cat("Genes removed:  ", nrow(dds) - sum(keep), "\n")
cat("Percent removed:", round(100 * (nrow(dds) - sum(keep)) / nrow(dds), 2), "%\n")

dds <- dds[keep, ]

processed_dds_group <- DESeq(dds)

resultsNames(processed_dds_group)

```


## PCA - global data check

```{r VST}
pca_path <- file.path(main_effects_path, "PCA")
make_subdir(pca_path)
```

## QC Global PCA  
```{r PCA check}
vsd_dds <- vst(processed_dds, blind = FALSE)

all_sampleIDs <- colnames(assay(vsd_dds))

# PCA 1: biology-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_colTime"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "cell_set", label = "genotype",
            save_prefix = save_prefix )

# PCA 2: QC-set-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_cell_set"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "cell_set", shape = "genotype", label = NA,
            save_prefix = save_prefix, scale_to_axes = TRUE)

# PCA 3: 
save_prefix <- file.path(pca_path, paste0(exp_name, "_time_geno"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "genotype", label = NA,
            save_prefix = save_prefix, scale_to_axes = TRUE)

```


```{r PCA check}
pcaLoadings(
  vsd_data = assay(vsd_dds),metadata = metadata, sample_ids = all_sampleIDs,
  exp_name = exp_name, name = "allSamples", results_dir = pca_path)

# print variance captured by each PCA 
p <- PCAtools::pca(assay(vsd_dds), metadata = metadata, removeVar = 0.9)
round(p$variance, 2)[1:6]
```


### PC2, 3, 4
```{r}

# PC1 vs PC3
p13_time <- PCAtools::biplot(p, x = "PC1", y = "PC3", 
                             colby = "cell_set", shape = "treatment", lab = metadata[["timepoint"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PCA_PC1_PC3_cell_set.png")),
       p13_time, width = 6, height = 5, dpi = 300)

# PC2 vs PC3
p23_time <- PCAtools::biplot(p, x = "PC2", y = "PC3",
                            colby = "cell_set", shape = "treatment", lab = metadata[["timepoint"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC2_PC3_cell_set.png")),
       p23_time, width = 6, height = 5, dpi = 300)


p23_time <- PCAtools::biplot(p, x = "PC2", y = "PC3",
                            colby = "timepoint", shape = "treatment", lab = metadata[["genotype"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC2_PC3_timepoint.png")),
       p23_time, width = 6, height = 5, dpi = 300)

# PC1 vs PC4
p14_batch <- PCAtools::biplot(p, x = "PC1", y = "PC4",
                             colby = "cell_set", shape = "treatment", lab = metadata[["timepoint"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC1_PC4_cell_set.png")),
       p14_batch, width = 6, height = 5, dpi = 300)

p14_batch <- PCAtools::biplot(p, x = "PC1", y = "PC4",
                             colby = "timepoint", shape = "cell_set", lab = metadata[["treatment"]], returnPlot = TRUE)
ggsave(file.path(pca_path, paste0(exp_name, "_PC1_PC4_tim.png")),
       p14_batch, width = 6, height = 5, dpi = 300)


```

### QC PCAs
```{r}
# D3 only
d3_ids <- rownames(
  metadata %>%
    dplyr::filter(timepoint == "D3")
)

vsd_d3 <- assay(vsd_dds)[, d3_ids]
meta_d3 <- metadata[d3_ids, , drop = FALSE]

generatePCA(vsd_d3, meta_d3, d3_ids,
            colby="cell_set", shape="genotype", label=NULL,
            save_prefix=file.path(pca_path, paste0(exp_name, "_QC_D3_only")),
            annotation_colors=annotation_colors)

# D3, D4, D6
early_ids <- rownames(
  metadata %>%
    dplyr::filter(timepoint %in% c("D3", "D4", "D6"))
)

vsd_early <- assay(vsd_dds)[, early_ids]
meta_early <- metadata[early_ids, , drop = FALSE]

generatePCA(vsd_early, meta_early, early_ids,
  colby = "timepoint", shape = "cell_set", label = "genotype",
  save_prefix = file.path(pca_path, paste0(exp_name, "_QC_D3_D4_D6_timepoint")),
  annotation_colors = annotation_colors)

generatePCA(vsd_early, meta_early, early_ids,
  colby = "treatment", shape = "cell_set", label = "genotype",
  save_prefix = file.path(pca_path, paste0(exp_name, "_QC_D3_D4_D6_treatment")),
  annotation_colors = annotation_colors
)

generatePCA(vsd_early, meta_early, early_ids,
  colby = "genotype", shape = "cell_set", label = "treatment",
  save_prefix = file.path(pca_path, paste0(exp_name, "_QC_D3_D4_D6_genotype")),
  annotation_colors = annotation_colors
)

```

### genotype specific PCAs
```{r}
# D21 only
d21_ids <- rownames(
  metadata %>%
    dplyr::filter(genotype == "D21")
)

vsd_d21 <- assay(vsd_dds)[, d21_ids]
meta_d21 <- metadata[d21_ids, , drop = FALSE]

generatePCA(vsd_d21, meta_d21, d21_ids,
            colby="timepoint", shape="treatment", label="cell_set",
            save_prefix=file.path(pca_path, paste0(exp_name, "_D21_only_time")),
            annotation_colors=annotation_colors)

generatePCA(vsd_d21, meta_d21, d21_ids,
            colby="cell_set", shape="treatment", label="timepoint",
            save_prefix=file.path(pca_path, paste0(exp_name, "_D21_only_cell")),
            annotation_colors=annotation_colors)


# T21 only
t21_ids <- rownames(
  metadata %>%
    dplyr::filter(genotype == "T21")
)

vsd_t21 <- assay(vsd_dds)[, t21_ids]
meta_t21 <- metadata[t21_ids, , drop = FALSE]

generatePCA(vsd_t21, meta_t21, t21_ids,
            colby="timepoint", shape="treatment", label="cell_set",
            save_prefix=file.path(pca_path, paste0(exp_name, "_T21_only_time")),
            annotation_colors=annotation_colors)

generatePCA(vsd_t21, meta_t21, t21_ids,
            colby="cell_set", shape="treatment", label="timepoint",
            save_prefix=file.path(pca_path, paste0(exp_name, "_T21_only_cell")),
            annotation_colors=annotation_colors)
```

### treatment specific PCAs
```{r}
# CTRL only
ctrl_ids <- rownames(
  metadata %>%
    dplyr::filter(treatment == "CTRL")
)

vsd_ctrl <- assay(vsd_dds)[, ctrl_ids]
meta_ctrl <- metadata[ctrl_ids, , drop = FALSE]

generatePCA(vsd_ctrl, meta_ctrl, ctrl_ids,
            colby="timepoint", shape="genotype", label="cell_set",
            save_prefix=file.path(pca_path, paste0(exp_name, "_CTRL_only_time")),
            annotation_colors=annotation_colors)

generatePCA(vsd_ctrl, meta_ctrl, ctrl_ids,
            colby="cell_set", shape="genotype", label="timepoint",
            save_prefix=file.path(pca_path, paste0(exp_name, "_CTRL_only_cell")),
            annotation_colors=annotation_colors)

# RA only
ra_ids <- rownames(
  metadata %>%
    dplyr::filter(treatment == "RA")
)

vsd_ra <- assay(vsd_dds)[, ra_ids]
meta_ra <- metadata[ra_ids, , drop = FALSE]


generatePCA(vsd_ra, meta_ra, ra_ids,
            colby="timepoint", shape="genotype", label="cell_set",
            save_prefix=file.path(pca_path, paste0(exp_name, "_RA_only_time")),
            annotation_colors=annotation_colors)

generatePCA(vsd_ra, meta_ra, ra_ids,
            colby="cell_set", shape="genotype", label="timepoint",
            save_prefix=file.path(pca_path, paste0(exp_name, "_RA_only_cell")),
            annotation_colors=annotation_colors)

generatePCA(vsd_ra, meta_ra, ra_ids,
            colby="genotype", shape="cell_set", label="timepoint",
            save_prefix=file.path(pca_path, paste0(exp_name, "_RA_only_geno")),
            annotation_colors=annotation_colors)

```


## Correlation heatmaps 
(using VST or batch-corrected)

### all samples
```{r}

heatmap_path <- file.path(main_effects_path, "Heatmap")
make_subdir(heatmap_path)


# SORT: genotype -> treatment -> timepoint -> cell_set
genotype_order <- factor(metadata$genotype, levels = c("D21", "T21"))
cell_set_order <- factor(metadata$cell_set, levels = c("Set3", "Set4"))

meta_hm <- metadata[
  order(genotype_order,
    metadata$treatment,
    metadata$timepoint,
    cell_set_order),
]

# subset VST matrix using defined order
vsd_all <- assay(vsd_dds)[, meta_hm$sampleID_unique, drop = FALSE]
cor_matrix <- cor(vsd_all)

rownames(cor_matrix) <- paste(
  meta_hm$genotype,
  meta_hm$timepoint,
  meta_hm$treatment,
  meta_hm$cell_set,
  sep = "_"
)

colnames(cor_matrix) <- meta_hm$sampleID_unique


col_annotation <- data.frame(
  timepoint = meta_hm$timepoint,
  treatment = meta_hm$treatment,
  genotype  = meta_hm$genotype,
  cell_set  = meta_hm$cell_set
)

rownames(col_annotation) <- meta_hm$sampleID_unique

# gap based on num of timepoints 
num_clusters <- length(unique(meta_hm$timepoint))

# generate heatmap 
heatmap_filename <- file.path(heatmap_path, paste0(exp_name, "_heatmap_allSamples.png"))
ph1 <- pheatmap(
  cor_matrix,
  filename = heatmap_filename,
  main = "Correlation Heatmap - All Samples",
  annotation_col = col_annotation,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  cutree_rows = num_clusters,
  display_numbers = FALSE,
  color = viridis::viridis(100),
  annotation_colors = annotation_colors,
  border_color = NA,
  fontsize_row = 10,
  fontsize_col = 8,
  height = 16,
  width = 10
)

```

### by treatment 
```{r}
unique_treatments <- unique(metadata$treatment)

# loop through each treatment and generate heatmap
for (trt in unique_treatments) {
  
  # sort: timepoint -> genotype -> cell_set
  meta_sub <- metadata[metadata$treatment == trt, ]
  
  genotype_order <- factor(meta_sub$genotype, levels = c("D21", "T21"))
  cell_set_order <- factor(meta_sub$cell_set, levels = c("Set3", "Set4"))
  
  meta_sub <- meta_sub[
    order(
      meta_sub$timepoint,
      genotype_order,
      cell_set_order
    ),
  ]
  
  vsd_subset <- assay(vsd_dds)[, meta_sub$sampleID_unique, drop = FALSE]
  cor_matrix <- cor(vsd_subset)
  
  rownames(cor_matrix) <- paste(
    meta_sub$genotype,
    meta_sub$timepoint,
    meta_sub$treatment,
    meta_sub$cell_set,
    sep = "_"
  )
  colnames(cor_matrix) <- meta_sub$sampleID_unique
  
  col_annotation <- data.frame(
    timepoint = meta_sub$timepoint,
    genotype  = meta_sub$genotype,
    cell_set  = meta_sub$cell_set
  )
  rownames(col_annotation) <- meta_sub$sampleID_unique
  
  num_clusters <- length(unique(meta_sub$timepoint))
  heatmap_filename <- file.path(heatmap_path, paste0(exp_name, "_heatmap_", trt, ".png"))
  
  phtrt <- pheatmap(
    cor_matrix,
    filename = heatmap_filename,
    main = paste("Correlation Heatmap -", trt, "Samples"),
    annotation_col = col_annotation,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    cutree_rows = num_clusters,
    display_numbers = FALSE,
    color = viridis::viridis(100),
    annotation_colors = annotation_colors,
    border_color = NA,
    fontsize_row = 10,
    fontsize_col = 8,
    height = 12,
    width = 10
  )
  
}



```


# -------------
# 1. DESeq2 compound effect 
```{r }
comp_effects_path <- file.path(save_dir, "Compound_Effects")
make_subdir(comp_effects_path)


dds_group <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ cell_set + genotype_timepoint_treatment
)

keep <- edgeR::filterByExpr(
  counts(dds_group),
  group = metadata$group)


cat("\n============================\n")
cat(" filterByExpr summary\n")
cat("============================\n")
cat("Total genes:    ", nrow(dds), "\n")
cat("Genes kept:     ", sum(keep), "\n")
cat("Genes removed:  ", nrow(dds) - sum(keep), "\n")
cat("Percent removed:", round(100 * (nrow(dds) - sum(keep)) / nrow(dds), 2), "%\n")

dds_group <- dds_group[keep, ]

processed_dds_group <- DESeq(dds_group)

resultsNames(processed_dds_group)

```

## QC Global PCA  
```{r PCA check}
pca_path <- file.path(comp_effects_path, "PCA")
make_subdir(pca_path)

vsd_dds <- vst(processed_dds_group, blind = FALSE)

all_sampleIDs <- colnames(assay(vsd_dds))

# PCA 1: biology-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_colTime"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "cell_set", label = "genotype",
            save_prefix = save_prefix )

# PCA 2: QC-set-first
save_prefix <- file.path(pca_path, paste0(exp_name, "_cell_set"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "cell_set", shape = "genotype", label = NA,
            save_prefix = save_prefix, scale_to_axes = TRUE)

# PCA 3: 
save_prefix <- file.path(pca_path, paste0(exp_name, "_time_geno"))
generatePCA(vsd = assay(vsd_dds), metadata = metadata, all_sampleIDs,
            colby = "timepoint", shape = "genotype", label = NA,
            save_prefix = save_prefix, scale_to_axes = TRUE)


# loadings
pcaLoadings(
  vsd_data = assay(vsd_dds),metadata = metadata, sample_ids = all_sampleIDs,
  exp_name = exp_name, name = "allSamples", results_dir = pca_path)

# print variance captured by each PCA 
p <- PCAtools::pca(assay(vsd_dds), metadata = metadata, removeVar = 0.9)
round(p$variance, 2)[1:6]
```


# 2. DEseq2 DEG analysis

```{r}
log2FC_Cutoff <- 0.5
fdrCutoff <- 0.05

deg_path <- file.path(comp_effects_path, paste0("L2FC", log2FC_Cutoff, "_FDR", fdrCutoff))
make_subdir(deg_path)

# Check metadata and DESeq object alignment
if (!all(rownames(metadata) == colnames(processed_dds_group))) {
  warning("⚠️ Rownames in metadata do NOT match colnames in processed_dds_group!")
  
  cat("🔍 Mismatch summary:\n")
  cat(" - Metadata rows:", nrow(metadata), "\n")
  cat(" - DESeq2 columns:", ncol(processed_dds_group), "\n")
  cat(" - First few metadata row names:\n", head(rownames(metadata)), "\n")
  cat(" - First few DESeq2 column names:\n", head(colnames(processed_dds_group)), "\n")
  
} else {
  cat("✅ Metadata rownames match DESeq2 column names.\n")
}

```


##  DE Analysis
-perform differential expression analysis for comparision across time points
-apply the DESeq2 package to extract differentially expressed genes (DEGs) and organizes in two list (a) filtered DEGS, up dn (b) unfilterd all genes 

```{r DEG comps}
# define DEG comparisons to make 
deg_comparisons <- list(
  "T21_CTRL_vs_D21_CTRL"   = list(genotype1 = "T21", genotype2 = "D21", treat1 = "CTRL", treat2 = "CTRL"),
  "T21_RA_vs_D21_RA"   = list(genotype1 = "T21", genotype2 = "D21", treat1 = "RA", treat2 = "RA"),
  "D21_RA_vs_D21_CTRL" = list(genotype = "D21", treat1 = "CTRL", treat2 = "RA"),
  "T21_RA_vs_T21_CTRL" = list(genotype = "T21", treat1 = "CTRL", treat2 = "RA")
)

timepoints <- unique(metadata$timepoint)

```


```{r DE analysis, message=FALSE}
# master lists 
filtered_results <- list()
unfiltered_results <- list()

for (comp in names(deg_comparisons)) {
  comp_info <- deg_comparisons[[comp]]
  
  # lists for current comparison
  filtered_list <- list()
  unfiltered_list <- list()
  
  for (d in timepoints) {
    
    # Dynamically build contrast groups and ID
    if (!is.null(comp_info$genotype1)) {
      x <- paste0(comp_info$genotype1, "_", d, "_", comp_info$treat1)
      y <- paste0(comp_info$genotype2, "_", d, "_", comp_info$treat2)
      id <- paste0(comp_info$genotype1, "_", d, "_", comp_info$treat1, 
                   "_vs_", comp_info$genotype2, "_", d, "_", comp_info$treat2)
    } else {
      genotype <- comp_info$genotype
      x <- paste0(genotype, "_", d, "_", comp_info$treat2)
      y <- paste0(genotype, "_", d, "_", comp_info$treat1)
      id <- paste0(genotype, "_", d, "_", comp_info$treat2, 
                   "_vs_", genotype, "_", d, "_", comp_info$treat1)
    }
    
    # Only run if both contrast groups exist
    if (x %in% unique(processed_dds_group$genotype_timepoint_treatment) && 
        y %in% unique(processed_dds_group$genotype_timepoint_treatment)) {
      
      res <- DESeq2::results(processed_dds_group, contrast = c("genotype_timepoint_treatment", x, y), alpha=fdrCutoff)
      res_shrunk <- lfcShrink(dds = processed_dds_group, contrast = c("genotype_timepoint_treatment", x, y), res = res, type = "ashr")
      
      # Add both raw and shrunk LFC using gene_id join
      res_df <- as.data.frame(res) %>% rownames_to_column("gene_id")
      res_shrunk_df <- as.data.frame(res_shrunk) %>% rownames_to_column("gene_id")
      
      # Join by gene_id and filter out rows with NA padj
      merged_df <- inner_join(res_shrunk_df, res_df[, c("gene_id", "log2FoldChange")], by = "gene_id") %>%
        filter(!is.na(padj)) %>%
        dplyr::rename(
          log2FoldChange_shrunk = log2FoldChange.x,
          log2FoldChange_raw = log2FoldChange.y
        ) %>%
        dplyr::select(gene_id, baseMean, log2FoldChange_shrunk, log2FoldChange_raw, pvalue, padj)
      
      unfiltered_list[[id]] <- merged_df

      # Filter significant DEGs (based on raw LFC)
      tmp <- merged_df %>%
        dplyr::filter(!is.na(padj), abs(log2FoldChange_raw) > log2FC_Cutoff, padj < fdrCutoff)

      filtered_list[[paste0(id, ".up")]] <- tmp %>% filter(log2FoldChange_raw > 0) %>% pull(gene_id)
      filtered_list[[paste0(id, ".dn")]] <- tmp %>% filter(log2FoldChange_raw < 0) %>% pull(gene_id)
    }
  }
  
  # Save per-comparison lists 
  saveRDS(filtered_list, file = file.path(rdata_path, paste0(exp_name, "_DEG_filtered_", comp, "_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".rds")))
  saveRDS(unfiltered_list, file = file.path(rdata_path, paste0(exp_name, "_DEG_unfiltered_", comp, ".rds")))
  
  # Export Excel of unfiltered data
  results_for_excel <- lapply(unfiltered_list, function(df) data.frame(Gene = df$gene_id, df[ , -1]))
  write_xlsx(results_for_excel, path = file.path(deg_path, paste0(exp_name, "_DEG_unfiltered_", comp, ".xlsx")))
  
  # Store into master result containers
  filtered_results[[comp]] <- filtered_list
  unfiltered_results[[comp]] <- unfiltered_list
}


saveRDS(filtered_results, file = file.path(rdata_path, paste0(exp_name, "_DEG_filtered_ALL_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".rds")))
saveRDS(unfiltered_results, file = file.path(rdata_path, paste0(exp_name, "_DEG_unfiltered_ALL.rds")))
```


## DEG Summary 
```{r}
deg_summary <- list()

for (comparison in names(filtered_results)) {
  contrast_list <- filtered_results[[comparison]]
  if (length(contrast_list) == 0) next
  
  for (contrast_id in names(contrast_list)) {
    direction <- ifelse(grepl("\\.up$", contrast_id), "Upregulated", "Downregulated")
    contrast_base <- sub("\\.(up|dn)$", "", contrast_id)
    deg_summary[[length(deg_summary) + 1]] <- data.frame(
      Comparison = comparison,
      Contrast   = contrast_base,
      Direction  = direction,
      NumGenes   = length(contrast_list[[contrast_id]])
    )
  }
}

deg_summary_df <- bind_rows(deg_summary) %>%
  arrange(Comparison, Contrast, Direction)

# Total DEG count across all contrasts
total_degs <- sum(deg_summary_df$NumGenes, na.rm = TRUE)

# CSV
save_csv(deg_summary_df, path = file.path(deg_path, paste0(exp_name, "_DEG_summary_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".csv")))

# PDF 
pdf_path <- file.path(deg_path, paste0(exp_name, "_DEG_summary_log2FC.", log2FC_Cutoff,
         "_FDR.", fdrCutoff, ".pdf"))

title_text <- paste0("DEG Summary | log2FC cutoff = ", log2FC_Cutoff,
                     " | FDR cutoff = ", fdrCutoff,
                     " | Total DEGs = ", total_degs)

pdf(pdf_path, width = 8.5, height = 11)
grid.text(title_text, gp = gpar(fontsize = 16, fontface = "bold"), x = 0.5, y = 0.97)  # centered top
gridExtra::grid.table(deg_summary_df)

dev.off()
```
#3.  SingScore 
Need: expression matrix, module gene lists
```{r}
library(singscore)

score_dir <- file.path(comp_effects_path, "Singscore")
make_subdir(score_dir)

expr_mat <- assay(vsd_dds)

# check overlap with expression matrix 
usable_scoring_gene_sets <- purrr::map(
  all_scoring_gene_sets,
  ~ intersect(.x, rownames(expr_mat))
)

usable_module_summary <- tibble::tibble(
  module = names(usable_scoring_gene_sets),
  n_genes = purrr::map_int(all_scoring_gene_sets, length),
  n_usable = purrr::map_int(usable_scoring_gene_sets, length)
)

usable_module_summary

# drop any empty modules 
usable_scoring_gene_sets <- usable_scoring_gene_sets[
  purrr::map_int(usable_scoring_gene_sets, length) > 0
]
```

## Run singscore
TotalScore →  actual module activity (what you care about)
TotalDispersion → variability of ranks within the gene set
```{r}
#rank expression matrix 
ranked_mat <- singscore::rankGenes(expr_mat)

#run singscore
score_list <- purrr::imap(
  usable_scoring_gene_sets,
  ~ {
    res <- singscore::simpleScore(
      rankData = ranked_mat,
      upSet = .x
    ) %>%
      as.data.frame() %>%
      tibble::rownames_to_column("sampleID_unique") %>%
      dplyr::select(sampleID_unique, TotalScore) %>%
      dplyr::rename(score = TotalScore) %>%
      dplyr::mutate(module = .y)
  }
)

score_df <- dplyr::bind_rows(score_list)

# add metadata
score_df <- score_df %>%
  dplyr::left_join(
    metadata,
    by = "sampleID_unique"
  ) %>%
  dplyr::mutate(day_num = as.numeric(gsub("D", "", timepoint)))

#check that scoring worked
dplyr::count(score_df, module)
sum(is.na(score_df$genotype)) # should be 0

# Save df

saveRDS(score_df,
  file.path(rdata_path, paste0(exp_name, "comp_singscore_scores.rds")))

readr::write_csv(score_df,
  file.path(score_dir, paste0(exp_name, "comp_singscore_scores.csv")))
```

### Plot scores 

Module function had this hard coded: 
      breaks = c(3, 4, 6, 8, 10),
      labels = c("D3", "D4", "D6", "D8", "D10")
```{r}
#module titles
module_titles <- list(
  g2m = "Cell Cycle (G2/M Progression)",
  e2f = "Cell Cycle (E2F Targets)",
  contractile = "Cardiomyocyte Contractile Function",
  sarcomere = "Sarcomere Organization",
  progenitor = "Cardiac Progenitor / Developmental Programing",
  junction = "Cell Junction & Structural Organization",
  canonical_ra = "Canonical Retinoic Acid Signaling",
  curated_ra_signaling = "RA Transcriptional Regulation",
  kegg_ra_metabolism = "Retinoic Acid Metabolism",
  ra_cellcycle = "RA-Regulated Cell Cycle Response",
  junction_organization = "Cell Junction & Structural Organization",
  xu_global_cm = "Cardiomyocyte Identity (Embryonic)",
  xu_atrial = "Atrial Cardiomyocyte Identity",
  xu_ventricular = "Ventricular Cardiomyocyte Identity",
  xu_shf_pharyngeal = "Second Heart Field / Pharyngeal Mesoderm",
  xu_anterior_lpm = "Anterior Lateral Plate Mesoderm",
  xu_san = "Sinoatrial Node Identity",
  xu_avc = "Atrioventricular Canal Identity",
  xu_offtarget = "Off-Target Lineage Signature",
  deBono_all_atrial = "Atrial Cardiomyocyte Identity (Refined)",
  deBono_a1_atrial = "Atrial Identity (Early / Progenitor-like)",
  deBono_a2_atrial = "Atrial Identity (Nodal-like)",
  deBono_a3_atrial = "Atrial Identity (Left Atrial)",
  deBono_a4_atrial = "Atrial Identity (Right Atrial)",
  deBono_ventricular = "Ventricular Cardiomyocyte Identity",
  deBono_epicardial = "Epicardial Identity",
  deBono_stromal = "Cardiac Stromal Identity",
  curated_pluripotency = "Pluripotency / Undifferentiated State (Curated)",
  curated_ra_metabolism = "Retinoic Acid Metabolism (Curated)",
  curated_hh_activity = "Hedgehog Signaling Activity (Curated)",
  curated_hh_feedback = "Hedgehog Feedback / Readout (Curated)",
  curated_wnt_activity = "Wnt Signaling Activity (Curated)",
  curated_wnt_output = "Canonical Wnt Signaling Output (Curated)",
  curated_RA_pSHF_target = "RA Cardiac-Specific Targets (Curated)"
)


# only valid modules are used
modules_to_plot <- intersect(
  names(module_titles),
  unique(score_df$module)
)

# loop and plot + save
purrr::walk(
  modules_to_plot,
  ~ plot_module_score(
      df = score_df,
      module_name = .x,
      title = module_titles[[.x]],
      output_dir = score_dir,
      file_prefix = paste0(exp_name, "_", .x),
      width = 6, height = 5,
      facet_by = "treatment"
    )
)
```

### Atrial vs ventricular plot 
```{r}
atrial_ventricular_df <- score_df %>%
  dplyr::filter(module %in% c("xu_atrial", "xu_ventricular")) %>%
  dplyr::select(sampleID_unique, module, score, genotype, treatment, timepoint, day_num) %>%
  tidyr::pivot_wider(
    names_from = module,
    values_from = score
  )

ggplot(
  atrial_ventricular_df,
  aes(x = xu_ventricular, y = xu_atrial, color = genotype)
) +
  geom_point(size = 3) +
  geom_path(aes(group = interaction(genotype, treatment)), alpha = 0.5) +
  facet_wrap(~ treatment) +
  scale_color_manual(values = annotation_colors$genotype) +
  labs(
    x = "Ventricular identity score",
    y = "Atrial identity score",
    color = NULL,
    title = "Atrial vs Ventricular Identity"
  ) +
  theme_pub() +
  theme(legend.position = "top")
```

```{r}
atrial_ventricular_df <- atrial_ventricular_df %>%
  dplyr::mutate(
    atrial_minus_ventricular = xu_atrial - xu_ventricular
  )

ggplot(
  atrial_ventricular_df,
  aes(x = day_num, y = atrial_minus_ventricular, color = genotype, group = genotype)
) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  stat_summary(fun = mean, geom = "point", size = 2.5) +
  facet_wrap(~ treatment) +
  scale_x_continuous(
    breaks = c(3, 4, 6, 8, 10),
    labels = c("D3", "D4", "D6", "D8", "D10")
  ) +
  scale_color_manual(values = annotation_colors$genotype) +
  labs(
    title = "Relative Atrial vs Ventricular Identity",
    x = NULL,
    y = "Atrial - Ventricular score",
    color = NULL
  ) +
  theme_pub() +
  theme(
    legend.position = "top",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```



### landscape plot
```{r}
# Cardiac Development vs Cell Cycle
landscape_dir <- file.path(score_dir, "Singscore_Landscape")
make_subdir(landscape_dir)

# define gene sets
cardiac_dev_genes <- usable_scoring_gene_sets[["progenitor"]]
cell_cycle_genes  <- usable_scoring_gene_sets[["e2f"]]


# rank genes for singscore
ranked_expr <- singscore::rankGenes(expr_mat)

# score each module
cardiac_dev_score <- singscore::simpleScore(
  ranked_expr,
  upSet = cardiac_dev_genes
)$TotalScore

cell_cycle_score <- singscore::simpleScore(
  ranked_expr,
  upSet = cell_cycle_genes
)$TotalScore

# combine scores with metadata
landscape_df <- tibble::tibble(
  sampleID_unique = colnames(expr_mat),
  Cardiac_Development = as.numeric(cardiac_dev_score),
  Cell_Cycle = as.numeric(cell_cycle_score)
) %>%
  dplyr::left_join(metadata, by = "sampleID_unique") %>%
  dplyr::mutate(
    timepoint = factor(timepoint, levels = c("D3", "D4", "D6", "D8", "D10")),
    treatment = factor(treatment, levels = c("CTRL", "RA")),
    genotype = factor(genotype, levels = c("D21", "T21"))
  )

# quick check
head(landscape_df)
```

#### all samples 
```{r}

# all samples

p_landscape_all <- ggplot(
  landscape_df,
  aes(
    x = Cardiac_Development,
    y = Cell_Cycle,
    color = timepoint,
    shape = treatment
  )) +
  geom_point(size = 4, alpha = 0.9) +
  scale_color_manual(values = annotation_colors$timepoint, drop = FALSE) +
  ggrepel::geom_text_repel(
  aes(label = genotype),
  size = 3,
  max.overlaps = Inf)+
  theme_pub(base_size = 14) +
  labs(
    title = "Cardiac development vs cell cycle landscape",
    x = "Cardiac Development Score",
    y = "Cell Cycle Score",
    color = "Timepoint",
    shape = "Treatment") +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top")

p_landscape_all

ggsave(file.path(landscape_dir, paste0(exp_name, "_allSamples_SingLandscape_CardiacDev_vs_CellCycle.pdf")),
  p_landscape_all, width = 5, height = 5, dpi = 300, bg = "white")
```


#### filtered
```{r}
# one treatment 
landscape_ctrl <- landscape_df %>%
  dplyr::filter(treatment == "RA")


p_landscape_ctrl <- ggplot(landscape_ctrl,
  aes(x = Cardiac_Development, y = Cell_Cycle,
    color = genotype, shape = timepoint)) +
  geom_point(size = 2, alpha = 0.8, stroke = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +
  scale_color_manual(values = annotation_colors$genotype, drop = FALSE) +
  theme_pub(base_size = 8) +
  labs(
    title = NULL,
    x = "Cardiac Development",
    y = "Cell Cycle",
    color = NULL, shape = NULL) +
  theme(
    legend.position = "top",
  )

p_landscape_ctrl

ggsave(file.path(landscape_dir, paste0(exp_name, "_RA_SingLandscape_CardiacDev_vs_CellCycle.pdf")), p_landscape_ctrl,
  width =42, height = 64,  units = "mm", dpi = 300, bg = "white") # for figure 
```

# 3. GSEA
```{r}

library(enrichplot)
library(msigdbr)

gsea_dir <- file.path(comp_effects_path, "GSEA")
make_subdir(gsea_dir)

#DEG data 
D21_unfiltered <- unfiltered_results[["D21_RA_vs_D21_CTRL"]] 
T21_unfiltered <-  unfiltered_results[["T21_RA_vs_T21_CTRL"]] 

#msigdbs
msig_h <- msigdbr(species = "Homo sapiens", collection = "H") # hallmark sets
msig_c2 <- msigdbr(species = "Homo sapiens", collection = "C2") # curated pathways 
msig_c5 <- msigdbr(species = "Homo sapiens", collection = "C5")  # GO term

# combine the relevant MSigDB collections
msig_gsea_db <- dplyr::bind_rows(msig_h, msig_c2, msig_c5)
```

### Cardiac 
```{r}
# terms to test
list_type <- "cardiac"

gsea_terms <- c(
  "GOBP_SARCOMERE_ORGANIZATION",
  "REACTOME_STRIATED_MUSCLE_CONTRACTION",
  "GOBP_CARDIAC_MUSCLE_CONTRACTION",
  "KEGG_CARDIAC_MUSCLE_CONTRACTION",
  "HALLMARK_G2M_CHECKPOINT",
  "HALLMARK_E2F_TARGETS" 
)

# Build TERM2GENE from selected MSigDB terms 
term2gene <- msig_gsea_db %>%
  dplyr::filter(gs_name %in% gsea_terms) %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::distinct()

# sanity check
setdiff(gsea_terms, unique(term2gene$gs_name))

# Define input contrasts
gsea_input_list <- list(
  D21_D6_RA_vs_CTRL = D21_unfiltered[["D21_D6_RA_vs_D21_D6_CTRL"]],
  D21_D8_RA_vs_CTRL = D21_unfiltered[["D21_D8_RA_vs_D21_D8_CTRL"]],
  T21_D6_RA_vs_CTRL = T21_unfiltered[["T21_D6_RA_vs_T21_D6_CTRL"]],
  T21_D8_RA_vs_CTRL = T21_unfiltered[["T21_D8_RA_vs_T21_D8_CTRL"]]
)
```

###  Run & save
```{r}

# Run clusterProfiler GSEA for each contrast
cp_gsea_list <- purrr::imap(
  gsea_input_list,
  ~ {
    ranks <- prep_gsea_ranks(.x)
    
    clusterProfiler::GSEA(
      geneList = ranks,
      TERM2GENE = term2gene,
      pvalueCutoff = 1,
      verbose = FALSE
    )
  }
)

names(cp_gsea_list)

# Save  GSEA results table
cp_gsea_results <- purrr::imap_dfr(
  cp_gsea_list,
  ~ {
    as_tibble(.x@result) %>%
      dplyr::mutate(contrast = .y)
  }
)

saveRDS(cp_gsea_results, file.path(rdata_path, paste0(exp_name, "_clusterProfiler_", list_type, "_GSEA_D6_D8_RA_vs_CTRL.rds")))

readr::write_csv(cp_gsea_results, file.path(gsea_dir, paste0(exp_name, "_clusterProfiler_", list_type, "_GSEA_D6_D8_RA_vs_CTRL.csv")))
```

### Plots
walk() applies a function (.f) only to each element, ignoring names or indices.
iwalk() passes both the element (.x) and its index/name (.y) to the function
```{r}
# Loop through all contrasts and pathways and save full plots
clean_file_label <- function(x) {
  gsub("[^A-Za-z0-9_]+", "_", x)
}

purrr::iwalk(
  cp_gsea_list,
  ~ {
    gsea_obj <- .x
    contrast_nm <- .y
    
    valid_ids <- intersect(gsea_terms, gsea_obj@result$ID)
    
    purrr::walk(
      valid_ids,
      function(pathway_nm) {
        
        res_row <- cp_gsea_results %>%
          dplyr::filter(
            contrast == contrast_nm,
            ID == pathway_nm
          )
        
        plot_title <- paste0(
          contrast_nm, "\n\n",
          pathway_nm, "\n",
          "NES = ", round(res_row$NES[1], 2),
          " | FDR = ", signif(res_row$p.adjust[1], 2)
        )
        
        p <- enrichplot::gseaplot2(
          gsea_obj,
          geneSetID = pathway_nm,
          title = plot_title,
          pvalue_table = FALSE
        )
        
        ggsave(filename = file.path(gsea_dir,
            paste0(clean_file_label(contrast_nm),
              "__", clean_file_label(pathway_nm), "_gseaplot.pdf")),
          plot = p, width = 6, height = 5, bg = "transparent"
        )
      }
    )
  }
)
```

###  NES heatmap summary of results
```{r}

cp_heatmap_df <- cp_gsea_results %>%
  dplyr::select(contrast, ID, NES) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = NES
  )

cp_heatmap_mat <- cp_heatmap_df %>%
  tibble::column_to_rownames("ID") %>%
  as.matrix()

cp_heatmap_mat <- cp_heatmap_mat[, c(
  "D21_D6_RA_vs_CTRL",
  "D21_D8_RA_vs_CTRL",
  "T21_D6_RA_vs_CTRL",
  "T21_D8_RA_vs_CTRL"
), drop = FALSE]

desired_order <- c(
  "GOBP_CARDIAC_MUSCLE_CONTRACTION",
  "GOBP_SARCOMERE_ORGANIZATION",
  "KEGG_CARDIAC_MUSCLE_CONTRACTION",
  "REACTOME_STRIATED_MUSCLE_CONTRACTION",
  "HALLMARK_E2F_TARGETS",
  "HALLMARK_G2M_CHECKPOINT"
)

cp_heatmap_mat <- cp_heatmap_mat[desired_order, , drop = FALSE]

annotation_col <- data.frame(
  genotype = c("D21", "D21", "T21", "T21"),
  timepoint = c("D6", "D8", "D6", "D8")
)

rownames(annotation_col) <- colnames(cp_heatmap_mat)
ph_annotation_colors <- annotation_colors
ph_annotation_colors$timepoint <- annotation_colors$timepoint[c("D6", "D8")]

nes_heatmap <- pheatmap::pheatmap(
  cp_heatmap_mat,
  color = colors,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ph_annotation_colors,
  main = "GSEA NES: RA vs CTRL at D6 and D8"
)

nes_heatmap

ggsave(filename = file.path(gsea_dir,
            paste0(exp_name,"_NESheatmap.pdf")),
          plot = nes_heatmap, width = 8, height = 5, bg = "transparent")
```


## DESeq RA interaction
```{r}
dds_int <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ cell_set + timepoint + genotype + treatment + genotype:treatment
)

dds_int <- DESeq(dds_int)
resultsNames(dds_int)
```




#XXXXXXXXXXXXXXXXXXX

# XXX 3. Chromosome Annotation
Use same bed file that was used to map reads to annotate chromosomes
- bed file give you chromosome -> tx_id 
- txi gives you tx_id ->gene_id 


## Chromsome Map 
### Read bed file - chromosome to tx_id mapping then tx_id to gene_id 
```{r}
chrom_dir <- file.path(save_dir, paste0("Chromosome Annotations"))
make_subdir(chrom_dir)

bed_cols <- c("chrom","chromStart","chromEnd","tx_id",
              "score","strand","thickStart","thickEnd",
              "itemRgb","blockCount","blockSizes","blockStarts")

genes_bed_raw <- readr::read_tsv(
  file.path(data_path, "genes.bed"),
  col_names = bed_cols,
  comment   = "#",
  show_col_types = FALSE
)

# normalize chromosome labels 
standard_chrs <- c(paste0("chr", 1:22), "chrX", "chrY")
norm_chr <- function(x) {
  x <- as.character(x)
  x <- ifelse(grepl("^chr", x), x, paste0("chr", x)) # add 'chr' if missing
  x
}

genes_bed <- genes_bed_raw %>%
  mutate(chrom = norm_chr(chrom)) %>%
  filter(chrom %in% standard_chrs) %>%
  distinct(tx_id, chrom)

# read tsv file - tx_id to gene_id mapping
tx2gene <- readr::read_tsv(
  file.path(star_salmon_path, "tx2gene.tsv"),
  col_names = c("tx_id","gene_id"),
  show_col_types = FALSE
) %>%
  distinct(tx_id, gene_id)


#check both 
cat("tx2gene rows:", nrow(tx2gene), 
    " unique tx:", dplyr::n_distinct(tx2gene$tx_id), "\n")
cat("tx in BED & tx2gene:", sum(tx2gene$tx_id %in% genes_bed$tx_id), "\n")
```

### Build final gene to chrom map
```{r}
gene_chrom <- genes_bed %>%
  dplyr::select(tx_id, chrom) %>%
  dplyr::left_join(tx2gene, by = "tx_id") %>%
  dplyr::select(gene_id, chrom) %>%
  dplyr::distinct()

saveRDS(gene_chrom,
  file = file.path(rdata_path, paste0(exp_name, "_gene_to_chrom.rds")))

#-- Check genes map to chromosome map --
# load DEGS
unfiltered_results <- readRDS( file.path(rdata_path, paste0(exp_name, "_DEG_unfiltered_ALL.rds")))
filtered_results <- readRDS( file.path(rdata_path, paste0(exp_name, "_DEG_filtered_ALL_log2FC.", log2FC_Cutoff, "_FDR.", fdrCutoff, ".rds")))

# full tested gene universe (unique, non-NA, character)
all_tested <- unique(na.omit(as.character(unlist(
  lapply(unfiltered_results, function(comp_list) {
    lapply(comp_list, function(df) df$gene_id)
  })
))))

# check that all genes mapped to chromosomes
all_degs <- unique(na.omit(as.character(unlist(filtered_results))))
mapped_tested <- length(intersect(all_tested, gene_chrom$gene_id))
mapped_degs   <- length(intersect(all_degs,   gene_chrom$gene_id))

cat(sprintf("Tested genes: %d | Mapped: %d (%.1f%%)\n",
          length(all_tested), mapped_tested, 100 * mapped_tested / max(1, length(all_tested))))
cat(sprintf("DEGs total:   %d | Mapped: %d (%.1f%%)\n",
          length(all_degs), mapped_degs, 100 * mapped_degs / max(1, length(all_degs))))
```

## Annotate DEG list 
```{r }
# Annotate DEG list 
chromosome_results <- lapply(filtered_results, function(day_list) {
  lapply(day_list, function(gene_vec) {
    tibble(gene_id = as.character(gene_vec)) %>%
      dplyr::left_join(gene_chrom, by = "gene_id")
  })
})

```


### Plot DEGS/ Chrom
```{r}
chrom_plots <- list()

# 1. summarize
chr_counts_df <- summarize_chr_counts(chromosome_results, 
                                      gene_chrom = gene_chrom, all_tested = all_tested )

# 2. clean up titles
chr_counts_df_pretty <- prettify_chr_counts(chr_counts_df)

# 3. plot with shared axis 
chrom_plots$facet_updown <- plot_chr_updown( chr_counts_df_pretty,
  save_dir = chrom_dir, mode = "facet")

# 4. up/down indiv graphs by day
chrom_plots$updown_day <- plot_chr_updown( chr_counts_df_pretty, save_dir = chrom_dir, mode = "individual")
```


## Total DEGs per chrosmome/day
```{r}

total_ratio_df <- summarize_total_ratio(chr_counts_df)
chrom_plots$total_byday <-plot_total_ratio_by_chrom_and_day(total_ratio_df, save_dir = chrom_dir, annotation_colors = annotation_colors)

```

## Save all
```{r}

saveRDS( chrom_plots,
  file = file.path(rdata_path, paste0(exp_name, "_chromosome_plots.rds")))

saveRDS(chr_counts_df_pretty, file = file.path(rdata_path, paste0(exp_name, "_chr_counts_pretty.rds")))

saveRDS(total_ratio_df, file = file.path(rdata_path, paste0(exp_name, "_total_ratio_df.rds")))

```






## GSEA
```{r}

library(enrichplot)
library(msigdbr)

gsea_dir <- file.path(comp_effects_path, "GSEA")
make_subdir(gsea_dir)

#DEG data 
D21_unfiltered <- unfiltered_results[["D21_RA_vs_D21_CTRL"]] 
T21_unfiltered <-  unfiltered_results[["T21_RA_vs_T21_CTRL"]] 

#msigdbs
msig_h <- msigdbr(species = "Homo sapiens", collection = "H") # hallmark sets
msig_c2 <- msigdbr(species = "Homo sapiens", collection = "C2") # curated pathways 
msig_c5 <- msigdbr(species = "Homo sapiens", collection = "C5")  # GO term

# combine the relevant MSigDB collections
msig_gsea_db <- dplyr::bind_rows(msig_h, msig_c2, msig_c5)
```

### Cardiac 
```{r}
# terms to test
list_type <- "cardiac"

gsea_terms <- c(
  "GOBP_SARCOMERE_ORGANIZATION",
  "REACTOME_STRIATED_MUSCLE_CONTRACTION",
  "GOBP_CARDIAC_MUSCLE_CONTRACTION",
  "KEGG_CARDIAC_MUSCLE_CONTRACTION",
  "HALLMARK_G2M_CHECKPOINT",
  "HALLMARK_E2F_TARGETS" 
)

# Build TERM2GENE from selected MSigDB terms 
term2gene <- msig_gsea_db %>%
  dplyr::filter(gs_name %in% gsea_terms) %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::distinct()

# sanity check
setdiff(gsea_terms, unique(term2gene$gs_name))

# Define input contrasts
gsea_input_list <- list(
  D21_D6_RA_vs_CTRL = D21_unfiltered[["D21_D6_RA_vs_D21_D6_CTRL"]],
  D21_D8_RA_vs_CTRL = D21_unfiltered[["D21_D8_RA_vs_D21_D8_CTRL"]],
  T21_D6_RA_vs_CTRL = T21_unfiltered[["T21_D6_RA_vs_T21_D6_CTRL"]],
  T21_D8_RA_vs_CTRL = T21_unfiltered[["T21_D8_RA_vs_T21_D8_CTRL"]]
)
```

###  Run & save
```{r}

# Run clusterProfiler GSEA for each contrast
cp_gsea_list <- purrr::imap(
  gsea_input_list,
  ~ {
    ranks <- prep_gsea_ranks(.x)
    
    clusterProfiler::GSEA(
      geneList = ranks,
      TERM2GENE = term2gene,
      pvalueCutoff = 1,
      verbose = FALSE
    )
  }
)

names(cp_gsea_list)

# Save  GSEA results table
cp_gsea_results <- purrr::imap_dfr(
  cp_gsea_list,
  ~ {
    as_tibble(.x@result) %>%
      dplyr::mutate(contrast = .y)
  }
)

saveRDS(cp_gsea_results, file.path(rdata_path, paste0(exp_name, "_clusterProfiler_", list_type, "_GSEA_D6_D8_RA_vs_CTRL.rds")))

readr::write_csv(cp_gsea_results, file.path(gsea_dir, paste0(exp_name, "_clusterProfiler_", list_type, "_GSEA_D6_D8_RA_vs_CTRL.csv")))
```

### Plots
walk() applies a function (.f) only to each element, ignoring names or indices.
iwalk() passes both the element (.x) and its index/name (.y) to the function
```{r}
# Loop through all contrasts and pathways and save full plots
clean_file_label <- function(x) {
  gsub("[^A-Za-z0-9_]+", "_", x)
}

purrr::iwalk(
  cp_gsea_list,
  ~ {
    gsea_obj <- .x
    contrast_nm <- .y
    
    valid_ids <- intersect(gsea_terms, gsea_obj@result$ID)
    
    purrr::walk(
      valid_ids,
      function(pathway_nm) {
        
        res_row <- cp_gsea_results %>%
          dplyr::filter(
            contrast == contrast_nm,
            ID == pathway_nm
          )
        
        plot_title <- paste0(
          contrast_nm, "\n\n",
          pathway_nm, "\n",
          "NES = ", round(res_row$NES[1], 2),
          " | FDR = ", signif(res_row$p.adjust[1], 2)
        )
        
        p <- enrichplot::gseaplot2(
          gsea_obj,
          geneSetID = pathway_nm,
          title = plot_title,
          pvalue_table = FALSE
        )
        
        ggsave(filename = file.path(gsea_dir,
            paste0(clean_file_label(contrast_nm),
              "__", clean_file_label(pathway_nm), "_gseaplot.pdf")),
          plot = p, width = 6, height = 5, bg = "transparent"
        )
      }
    )
  }
)
```

###  NES heatmap summary of results
```{r}

cp_heatmap_df <- cp_gsea_results %>%
  dplyr::select(contrast, ID, NES) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = NES
  )

cp_heatmap_mat <- cp_heatmap_df %>%
  tibble::column_to_rownames("ID") %>%
  as.matrix()

cp_heatmap_mat <- cp_heatmap_mat[, c(
  "D21_D6_RA_vs_CTRL",
  "D21_D8_RA_vs_CTRL",
  "T21_D6_RA_vs_CTRL",
  "T21_D8_RA_vs_CTRL"
), drop = FALSE]

desired_order <- c(
  "GOBP_CARDIAC_MUSCLE_CONTRACTION",
  "GOBP_SARCOMERE_ORGANIZATION",
  "KEGG_CARDIAC_MUSCLE_CONTRACTION",
  "REACTOME_STRIATED_MUSCLE_CONTRACTION",
  "HALLMARK_E2F_TARGETS",
  "HALLMARK_G2M_CHECKPOINT"
)

cp_heatmap_mat <- cp_heatmap_mat[desired_order, , drop = FALSE]

annotation_col <- data.frame(
  genotype = c("D21", "D21", "T21", "T21"),
  timepoint = c("D6", "D8", "D6", "D8")
)

rownames(annotation_col) <- colnames(cp_heatmap_mat)
ph_annotation_colors <- annotation_colors
ph_annotation_colors$timepoint <- annotation_colors$timepoint[c("D6", "D8")]

nes_heatmap <- pheatmap::pheatmap(
  cp_heatmap_mat,
  color = colors,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ph_annotation_colors,
  main = "GSEA NES: RA vs CTRL at D6 and D8"
)

nes_heatmap

ggsave(filename = file.path(gsea_dir,
            paste0(exp_name,"_NESheatmap.pdf")),
          plot = nes_heatmap, width = 8, height = 5, bg = "transparent")
```


## DESeq RA interaction
```{r}
dds_int <- DESeqDataSetFromTximport(
  txi = txi,
  colData = metadata,
  design = ~ cell_set + timepoint + genotype + treatment + genotype:treatment
)

dds_int <- DESeq(dds_int)
resultsNames(dds_int)
```

