---
title: "01: Normalize and Combine All Experiments"
date: "2026-02-02"
params:                              
  experiment_id: S3_COMB
  author: "Shalise Burch"
output: html_document  
knitr:
  opts_chunk:
    echo: TRUE        
    warning: FALSE     
    message: FALSE  
---

COMBINED FLOW DATA FOR SET3

## **SETUP**
##### Load Libraries
```{r libraries, message=FALSE, warning=FALSE}
# UPDATED:
# 2026-02-09 14:41 SAB ------------------------------

library(here)
library(readxl)
library(dplyr)
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(stringr) 
library(janitor)

```

##### Define Variables
```{r define pathes}
exp_name <- params$experiment_id
data_path <- here("flow", "01_data")
results_path <- here("flow", "02_results")
rdata_path <- here("flow", "03_rdata")

source(here("flow", "00_scripts", "flow_functions.R"))
```


```{r define customizations}
color_map <- data.frame(
  Group = c("CTRL", "SAG", "VIS", "RA", "RA_SAG", "RA_VIS"),
  Color = c("#999999", "#0072B2", "#E69F00", "#D55E00", "#56B4E9", "#009E73")
)

treatment_order <- color_map$Group

measurements <- tibble(
  measurement = c("dbl_neg", "isl_pos", "islctnt_pos", "ctnt_pos", "tot_isl", "tot_ctnt"),
  measurement_label = c("Double Negative", "Islet+", "Islet+ and cTnT+", "cTnT+", "Total Islet1+", "Total cTnT+")
)

#save to load and use later
save(
  color_map, treatment_order, measurements, 
  file = file.path(rdata_path, paste0(exp_name, "_customization_data.RData")))
```



## STEP1: Load  and Inspect Data
```{r import and inspect}
SAB01 <- read.csv(file.path(data_path, "SAB01_flow_dataFinal.csv"))
SAB04 <- read.csv(file.path(data_path, "SAB04_flow_dataFinal.csv"))
SAB08 <- read.csv(file.path(data_path, "SAB08_flow_dataFinal.csv"))

SAB01
SAB04
SAB08
```


```{r import and inspect}
# ISLET expression by day and treatment
ggboxplot(data, x = "day", y = "tot_isl", color = "cell", ylab = "Expression", xlab = "Day") + 
  facet_grid(. ~ treatment, labeller = label_both)

# CTNT expression by day and treatment
ggboxplot(data, x = "day", y = "tot_ctnt", color = "cell", ylab = "Expression", xlab = "Day") + 
  facet_grid(. ~ treatment, labeller = label_both)

```

## STEP 2: Clean Data

##### Transform Data 
Check your data before running *** 
Only run % change if percentages were saved as decimals in Excel (e.g., 0.45 instead of 45)
```{r convert decimals to percent values}

data <- data %>%
  mutate(
    across(all_of(measurements$measurement), ~ .x * 100),  # convert decimal % to actual %
    treatment = toupper(treatment),                            
    day_lb = toupper(day_lb),
    cell = toupper(cell),
    set = str_extract(cell, "^S\\d"),                    # extract "S2", "S3"
    genotype = str_extract(cell, "D21|T21"),             # extract "D21" or "T21"
    genotype = factor(genotype, levels = c("D21", "T21")),  # enforce order
    set = factor(set, levels = unique(set))
  )

data
```

## STEP 3: Reshape Data for Plots
Pivot data to long format for ggplot and statistical analysis
```{r long form}

long_data <- data %>%
  pivot_longer(
    cols = all_of(measurements$measurement), 
    names_to = "measurement",
    values_to = "value"
  ) %>%
 left_join(measurements, by = "measurement") %>%
  mutate(  
     # Factor  for  order in graph
    measurement = factor(measurement, levels = measurements$measurement),        
    measurement_label = factor(
      measurement_label,
      levels = measurements$measurement_label[match(measurements$measurement, measurements$measurement)]),
    treatment = factor(treatment, levels = treatment_order),
    genotype = factor(genotype, levels = c("D21", "T21")),
    set = factor(set, levels = unique(set))
  )

long_data
```


## STEP 4: Summary Statistics
##### Produce summary stats for each measurment
```{r sumstats}
measurement_names <- measurements$measurement

results_list <- list() 

for (m in measurement_names) {
  summary_data <- data %>%
    group_by(cell, day, treatment) %>%
    summarise(
      n    = sum(!is.na(.data[[m]])),        # count non-NA values
      mean = mean(.data[[m]], na.rm = TRUE),
      sd   = sd(.data[[m]], na.rm = TRUE),
      sem  = sd / sqrt(n), 
      .groups = "drop"
    ) %>%
    mutate(measurement = m)  # track which measurement it is
  
  results_list[[m]] <- summary_data
}

# Create lookup table from long_data
treatment_lookup <- long_data %>%
  select(treatment, treatment_label) %>%
  distinct()

# After binding all measurements together
sumstats_all <- bind_rows(results_list) %>%
  left_join(treatment_lookup, by = "treatment") %>%
  mutate(
    genotype = str_extract(cell, "D21|T21"),
    set = str_extract(cell, "^S\\d")
  ) %>%
  left_join(measurements, by = "measurement") %>%
  mutate(
    measurement_label = factor(measurement_label, levels = measurements$measurement_label),
    measurement       = factor(measurement, levels = measurements$measurement),
    treatment         = factor(treatment, levels = treatment_order),
    genotype          = factor(genotype, levels = c("D21", "T21")),
    set               = factor(set, levels = unique(set)),
    treatment_label   = factor(treatment_label, levels = unique(treatment_lookup$treatment_label)) # optional: keep order consistent
  )

print(sumstats_all)
```

##### SAVE DATA
```{r save data}

save_csv(data,        path = file.path(data_path, paste0(exp_name, "_flow_dataFinal.csv")))
save_csv(long_data,   path = file.path(data_path, paste0(exp_name, "_flow_longdata.csv")))
save_csv(sumstats_all, path = file.path(data_path, paste0(exp_name, "_flow_sumstats.csv")))

save(data, long_data, sumstats_all, 
  file = file.path(rdata_path, paste0(exp_name, "_flow_data.RData")))

```

#### Plot with all measurments
```{r all measurments }

bargraph.all<- ggplot(sumstats_all, aes(x = interaction(day), y = mean, fill = cell)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, position = position_dodge(0.9)) +
  facet_grid(measurement_label ~ treatment, scales = "free_y") + 
  labs(title = paste0("All Flow Measurements - ", exp_name),
       x = "Day",
       y = NULL,
       fill = "Cell Type") +
  theme_pub() +
  theme(panel.grid.major = element_line(colour = "#fafafa"))

ggsave( filename = paste0(exp_name, "_all_measurements.pdf"),
  plot = bargraph.all, path = results_path,
  height = 11, width = 10, units = "in", bg = "transparent")

```