---
title: "02: Assumptions and Statistical Comparisons -- D21 vs T21 "
params:                              
  experiment_id: SAB01
  author: "Shalise Burch"
  date: "2024-04-10"
output: html_document  
knitr:
  opts_chunk:
    echo: TRUE        
    warning: FALSE     
    message: FALSE   
---

# Analysis overview
For each day and each measurement:
    -> Compare T21 vs D21 within each treatment 

** 🔁 Statistical Comparison Workflow – Step-by-Step **
STEP1: Load Data and customization
STEP2: Run Statistical Analysis  
  a) check assumption 
  - Normality (Shapiro-Wilk)
  - Homogeneity of variance (Levene’s)
  b) Statistical Analysis  
  - t-test or Wilcoxon depending on assumptions
STEP4: Generate Bar Plots w/ Annotations


STEP1: Load Libraries and Data 
```{r loading libraries, include=FALSE}
# UPDATED:
# 2026-01-25 19:06 SAB ------------------------------

library(dplyr)
library(tidyverse)
library(readr)
library(ggpubr)
library(car)
library(here)
library(janitor)

genotype1 <- "D21"
genotype2 <- "T21"
exp_name <- params$experiment_id

data_path <- here("flow", "01_data")
results_path <- here("flow", "02_results")
rdata_path <- here("flow", "03_rdata")

source(here("flow", "00_scripts", "flow_functions.R"))

save_path <- file.path(results_path, "D21 vs T21")
make_subdir(save_path)

# load data
load(file.path(rdata_path, paste0(exp_name,  "_customization_data.RData")))
load(file.path(rdata_path, paste0(exp_name, "_flow_data.RData")))

```

## ---------------------
## Statistical Analysis
#### Assumptions and then stats 
```{r genotype stats}
pairwise_stats <- run_pairwise_stats(data = long_data,
  measurements = measurements, group_var = "genotype", 
  group1 = genotype1, group2 = genotype2,
  stratify_var = "treatment", id_var = "exp_no")

save_csv(data = pairwise_stats, path = file.path(save_path, paste0(exp_name, "_stats_D21vT21.csv")))

```


## -----------------------------
## PLOTS with stat annotation

```{r subfolder }
plots_folder <- file.path(save_path, "Plots")
make_subdir(plots_folder)

plot_recipes <- list()

for (measure in measurements$measurement) {
  for (treat_val in unique(long_data$treatment)) {
    treat_val <- as.character(treat_val)

    p <- make_flow_plot(
      measurement_name = measure, summary_data = sumstats_all,
      raw_data = long_data, treatment_val = treat_val,
      folder = save_path, exp_name = exp_name, color_map = color_map)
    
    #remove legend labels 
    p <- p +
      scale_fill_manual(name = "D21", values = setNames(color_map$Color, color_map$Group), labels = c("")) +
    scale_color_manual(name = "T21", values = setNames(color_map$Color, color_map$Group),  labels = c("")) 
    
    # Annotate stats — comparing genotype *within* each treatment
    p_annot <- add_pairwise_annotation(
      plot = p, stats_data = pairwise_stats,           
      measurement = measure,
      group_col = "genotype",                
      stratify_by = "treatment",             
      stratify_level = treat_val,
      show_pval = TRUE, show_effect = TRUE)

    ggsave(filename = paste0(exp_name, "_", measure, "_", treat_val, "_stats.pdf"),
      plot = p_annot, path = plots_folder,
      width = 5, height = 7, units = "in"
    )
    
    plot_key <- paste0(measure, "_", treat_val)
    
    plot_recipes[[plot_key]] <- list(
      measurement_name = measure,
      treatment_val = treat_val,
      summary_data = sumstats_all,
      raw_data = long_data,
      folder = save_path, exp_name = exp_name,
      color_map = color_map)
    
  }
}

saveRDS(plot_recipes, file = file.path(rdata_path, paste0(exp_name, "_D21vsT21_plot_recipes.rds")))



```


##### CUSTOM ------ 
```{r}
days <- c(3,4,6,8,10)
sumstats_filter  <- sumstats_all %>% filter(day %in% days)
long_data_filter <- long_data %>% filter(day %in% days)

plots_folder <- file.path(save_path, "Plots - D3,D4,D6,D8,D10")
make_subdir(plots_folder)
```

### Generate plots 
```{r}

plot_recipes <- list()

for (measure in measurements$measurement) {
  for (treat_val in unique(long_data_filter$treatment)) {
    treat_val <- as.character(treat_val)

    p <- make_flow_plot(
      measurement_name = measure, summary_data = sumstats_filter,
      raw_data = long_data_filter, treatment_val = treat_val,
      folder = save_path, exp_name = exp_name, color_map = color_map)
    
    #remove legend labels 
    p <- p +
      scale_fill_manual(name = "D21", values = setNames(color_map$Color, color_map$Group), labels = c("")) +
    scale_color_manual(name = "T21", values = setNames(color_map$Color, color_map$Group),  labels = c("")) 
    
    # Annotate stats — comparing genotype *within* each treatment
    p_annot <- add_pairwise_annotation(
      plot = p, stats_data = pairwise_stats,           
      measurement = measure,
      group_col = "genotype",                
      stratify_by = "treatment",             
      stratify_level = treat_val,
      show_pval = TRUE, show_effect = TRUE)

    ggsave(filename = paste0(exp_name, "_", measure, "_", treat_val, "_stats.pdf"),
      plot = p_annot, path = plots_folder,
      width = 5, height = 7, units = "in"
    )
    
    plot_key <- paste0(measure, "_", treat_val)
    
    plot_recipes[[plot_key]] <- list(
      measurement_name = measure,
      treatment_val = treat_val,
      summary_data = sumstats_filter,
      raw_data = long_data_filter,
      folder = save_path, exp_name = exp_name,
      color_map = color_map)
    
  }
}

saveRDS(plot_recipes, file = file.path(rdata_path, paste0(exp_name, "_D21vsT21_plot_recipesD346810.rds")))


```





