---
title: "02: Assumptions and Statistical Comparisons -- D21 vs T21 "
date: "2024-04-10"
params:                              
  experiment_id: SAB01
  author: "Shalise Burch"
output: html_document  
knitr:
  opts_chunk:
    echo: TRUE        
    warning: FALSE     
    message: FALSE   
---

# Analysis overview
For each day and each measurement:
    -> Compare CTRL vs RA within each genotype 

** 🔁 Statistical Comparison Workflow – Step-by-Step **
STEP1: Setup
  a) Load Libraries
  b) Set Variables 
  c) Load Data
STEP2: Run Statistical Analysis  
  a) check assumption 
  - Normality (Shapiro-Wilk)
  - Homogeneity of variance (Levene’s)
  b) Statistical Analysis  
  - t-test or Wilcoxon depending on assumptions
STEP4: Generate Bar Plots w/ Annotations


## Setup 
```{r loading libraries, include=FALSE}
# UPDATED:
# 2026-01-25 19:06 SAB ------------------------------

library(dplyr)
library(tidyverse)
library(readr)
library(ggpubr)
library(car)
library(here)
library(janitor)

exp_name <- params$experiment_id
data_path <- here("flow", "01_data")
results_path <- here("flow", "02_results")
rdata_path <- here("flow", "03_rdata")

source(here("flow", "00_scripts", "flow_functions.R"))

load(file.path(rdata_path, paste0(exp_name, "_customization_data.RData")))
load(file.path(rdata_path, paste0(exp_name, "_flow_data.RData")))

```


### Filter data if needed
```{r filter data}
group1 <- "CTRL"
group2 <- "RA"
comparison <- paste0(group1, "_vs_", group2)

save_path <- file.path(results_path, comparison)
make_subdir(save_path)

filtered_data <- long_data %>%
  filter(treatment %in% c(group1, group2)) %>%
  droplevels()

filtered_summary <- sumstats_all %>%
  filter(treatment %in% c(group1, group2)) %>%
  droplevels()

```

### filter days if needed 
```{r}
treatments_of_interest <- c(group1, group2)

valid_days <- long_data %>%
  filter(treatment %in% treatments_of_interest) %>%
  distinct(day, treatment) %>%
  count(day) %>%
  filter(n == length(treatments_of_interest)) %>%
  pull(day)

message("Overlapping days for ", group1, " and ", group2, ": ", paste(valid_days, collapse = ", "))

filtered_summary  <- filtered_summary %>% filter(day %in% valid_days)
filtered_data <- filtered_data %>% filter(day %in% valid_days)
```


## Statistical Analysis
```{r treatment stats}
pairwise_stats <- run_pairwise_stats(
  data = filtered_data, measurements = measurements,         
  group_var = "treatment", group1 = group1, group2 = group2,
  stratify_var = "genotype",id_var = "exp_no")

save_csv(data = pairwise_stats, path = file.path(save_path, paste0(exp_name,  "_", comparison, "_stats.csv")))

```


## Plots with stat annotation
```{r}
plots_folder <- file.path(save_path, "Plots")
make_subdir(plots_folder)

filtered_summary <- filtered_summary %>% filter(day != 3)
filtered_data <- filtered_data %>% filter(day != 3)

plot_recipes <- list()

for (measure in measurements$measurement) {
  for (cell_val in unique(filtered_data$genotype)) {
    cell_val <- as.character(cell_val)

    p <- make_flow_plot(
      measurement_name = measure,
      summary_data = filtered_summary,
      raw_data = filtered_data,
      cell_val = cell_val,
      folder = save_path,
      exp_name = exp_name,
      color_map = color_map
    )
    
    if (grepl("D21", cell_val)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
          color = "none", alpha = "none")
    } else if (grepl("T21", cell_val)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = "none",
          color = guide_legend(order = 1, override.aes = list(fill = NA)),
          alpha = "none"
        )
    }
    
    # Annotate stats — comparing treatment *within* each genotype
    p_annot <- add_pairwise_annotation(
      plot = p,
      stats_data = pairwise_stats,           
      measurement = measure,
      group_col = "treatment",                
      stratify_by = "genotype",             
      stratify_level = cell_val,
      show_pval = TRUE, 
      show_effect = TRUE
    )

    # Save it
    ggsave(filename = paste0(exp_name, "_", measure, "_", cell_val, "_", comparison, "_stats.pdf"),
      plot = p_annot, path = plots_folder,
      width = 5, height = 7, units = "in"
    )
    
    plot_key <- paste0(measure, "_", cell_val)
    plot_recipes[[plot_key]] <- list(
      measurement_name = measure,
      summary_data = filtered_summary,
      raw_data = filtered_data, cell_val = cell_val,
      folder = save_path, exp_name = exp_name,
      color_map = color_map,
      group_col = "treatment",
      stratify_by = "genotype",
      stratify_level = cell_val,
      show_pval = TRUE, show_effect = TRUE
    )

  }
}

saveRDS(plot_recipes, file = file.path(rdata_path, paste0(exp_name, "_", comparison, "_plot_recipes.rds")))


```

