---
title: "03: Assumptions and Statistical Comparisons -- selected treatments"
experiment_ID: "SAB04"
date: "2024-11-11"
output: 
  html_document:
    code_folding: "hide"  
knitr:
  opts_chunk:
    echo: TRUE        
    warning: FALSE    
    message: FALSE     
    fig.width: 6       
    fig.height: 4      
    chunk_labels: TRUE 

---
#### Last updated: 04-12-25 

Analysis: 
1. CTRL, SAG, VIS 
2. CTRL, RA, RA + VIS
3. RA, RA +SAG, RA + VIS


** Statistical Comparison Workflow – Step-by-Step **
STEP 1: Setup 
  a) Load Libraries
  b) Set Variables 
  c) Load Data
STEP 2: Run Statistical Analysis  
  a) Check assumption 
  - Normality (Shapiro-Wilk)
  - Homogeneity of variance (Levene’s)
  b) Statistical Analysis  
  - t-test or Wilcoxon depending on assumptions
STEP 3: Generate Bar Plots w/ Annotations
  a) Build plot 
  b) add annotations



## STEP 1: Setup

##### - Load libraries
```{r loading libraries, include=FALSE}

library(dplyr)
library(tidyverse)
library(readr)
library(ggpubr)
library(car)
library(here)
library(janitor)

```

##### - Define Variable 
```{r variables}
exp_name <- "SAB04"

data_path <- here("flow", "01_data")
results_path <- here("flow", "02_results")
rdata_path <- here("flow", "03_rdata")

source(here("flow", "00_scripts", "flow_functions.R"))

```

##### - Load Data  
```{r data}

load(file.path(rdata_path, paste0(exp_name, "_customization_data.RData")))
load(file.path(rdata_path, paste0(exp_name, "_flow_data.RData")))

```

# --------------------
# 1. CTRL, SAG, VIS 
# --------------------

#### - Filter data
```{r}
comparision <- "CTRL_SAG_VIS"

save_path <- file.path(results_path, comparision)
make_subdir(save_path)

filtered_data <- long_data %>%
  filter(treatment %in% c("CTRL", "SAG", "VIS")) %>%
  droplevels()

filtered_summary <- sumstats_all %>%
  filter(treatment %in% c("CTRL", "SAG", "VIS")) %>%
  droplevels()

```

## STEP 2:  Statistical Analysis

#### - Assumptions and Stats 
```{r treatment }
multi_results <- multi_flow_test(
  data = filtered_data,
  measurements = measurements,
  group_var = "treatment",
  subject_id = "exp_no",
  stratify_var = "genotype",       # compare treatments within each genotype
  day_var = "day",
  value_var = "value",
  save_dir = save_path,
  comparison = paste0(exp_name, "_", comparision)
)

saveRDS(multi_results, file = file.path(rdata_path, paste0(exp_name, "_", comparision, "_stat_results.rds")))
```



## STEP 3 : PLOTS with stat annotation

```{r subfolder }
plots_folder <- file.path(save_path, "Plots")
make_subdir(plots_folder)
```

##### - Generate plots with annotaions
```{r}
filtered_summary <- filtered_summary %>% filter(day != 3)
filtered_raw <- filtered_data %>% filter(day != 3)

plot_list <- list()

for (measure in measurements$measurement) {
  for (geno in unique(filtered_data$genotype)) {
    geno <- as.character(geno)
    # Create the plot
    p <- make_flow_plot(
      measurement_name = measure,
      summary_data = filtered_summary,
      raw_data = filtered_raw,
      cell_val = geno,  
      folder = save_path,
      exp_name = exp_name,
      color_map = color_map
    )
    
        # 🪄 Override guides based on cell type
    if (grepl("D21", geno)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
          color = "none", 
          alpha = "none"
        )
    } else if (grepl("T21", geno)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = "none",
          color = guide_legend(order = 1, override.aes = list(fill = NA)),
          alpha = "none"
        )
    }

    # Annotate stats 
    p_annot <- add_multi_annotation(
      plot = p,
      stats_data = multi_results,           
      measurement = measure,
      group_col = "treatment",                
      stratify_by = "genotype",             
      stratify_level = geno,
      show_effect = FALSE,
      show_pval = FALSE
    )
    
    p_annot <- add_friedman_annotation(
      plot = p_annot,
      stats_data = multi_results,
      measurement = measure,
      stratify_by = "genotype",
      stratify_level = geno
    )

    # Save it
    ggsave(filename = paste0(exp_name, "_", measure, "_", geno, "_stats.pdf"),
      plot = p_annot, path = plots_folder,
      width = 6, height = 7, units = "in"
    )
    
    plot_key <- paste0(measure, "_", geno)
    plot_list[[plot_key]] <- p_annot
  }
}

saveRDS(plot_list, file = file.path(rdata_path, paste0(exp_name, "_alltreatments_plots.rds")))


```

# --------------------
# 2. CTRL, RA, RA + VIS
# --------------------

```{r}
comparision <- "CTRL_RA_RAVIS"

save_path <- file.path(results_path, comparision)
make_subdir(save_path)

filtered_data <- long_data %>%
  filter(treatment %in% c("CTRL", "RA", "RA_VIS")) %>%
  droplevels()

filtered_summary <- sumstats_all %>%
  filter(treatment %in% c("CTRL", "RA", "RA_VIS")) %>%
  droplevels()

```

## STEP 2:  Statistical Analysis

#### - Assumptions and Stats 
```{r treatment }

multi_results <- multi_flow_test(
  data = filtered_data,
  measurements = measurements,
  group_var = "treatment",
  subject_id = "exp_no",
  stratify_var = "genotype",       # compare treatments within each genotype
  day_var = "day",
  value_var = "value",
  save_dir = save_path,
  comparison = paste0(exp_name, "_", comparision)
)

saveRDS(multi_results, file = file.path(rdata_path, paste0(exp_name, "_", comparision, "_stat_results.rds")))
```



## STEP 3 : PLOTS with stat annotation

```{r subfolder }

plots_folder <- file.path(save_path, "Plots")
make_subdir(plots_folder)
```

##### - Generate plots with annotaions
```{r}

filtered_summary <- filtered_summary %>% filter(day != 3)
filtered_raw <- filtered_data %>% filter(day != 3)

plot_list <- list()

for (measure in measurements$measurement) {
  for (geno in unique(filtered_data$genotype)) {
    geno <- as.character(geno)
    # Create the plot
    p <- make_flow_plot(
      measurement_name = measure,
      summary_data = filtered_summary,
      raw_data = filtered_raw,
      cell_val = geno,  
      folder = save_path,
      exp_name = exp_name,
      color_map = color_map
    )
    
        # 🪄 Override guides based on cell type
    if (grepl("D21", geno)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
          color = "none", 
          alpha = "none"
        )
    } else if (grepl("T21", geno)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = "none",
          color = guide_legend(order = 1, override.aes = list(fill = NA)),
          alpha = "none"
        )
    }

    # Annotate stats 
    p_annot <- add_multi_annotation(
      plot = p,
      stats_data = multi_results,           
      measurement = measure,
      group_col = "treatment",                
      stratify_by = "genotype",             
      stratify_level = geno,
      show_effect = FALSE,
      show_pval = FALSE
    )
    
    p_annot <- add_friedman_annotation(
      plot = p_annot,
      stats_data = multi_results,
      measurement = measure,
      stratify_by = "genotype",
      stratify_level = geno
    )

    # Save it
    ggsave(filename = paste0(exp_name, "_", measure, "_", geno, "_stats.pdf"),
      plot = p_annot, path = plots_folder,
      width = 6, height = 7, units = "in"
    )
    
    plot_key <- paste0(measure, "_", geno)
    plot_list[[plot_key]] <- p_annot
  }
}

saveRDS(plot_list, file = file.path(rdata_path, paste0(exp_name, "_alltreatments_plots.rds")))


```


# --------------------
# 3. RA, RA+SAG, RA+VIS
# --------------------

```{r}

comparision <- "RA_RASAG_RAVIS"

save_path <- file.path(results_path, comparision)
make_subdir(save_path)

filtered_data <- long_data %>%
  filter(treatment %in% c("RA", "RA_SAG", "RA_VIS")) %>%
  droplevels()

filtered_summary <- sumstats_all %>%
  filter(treatment %in% c("RA", "RA_SAG", "RA_VIS")) %>%
  droplevels()

```

## STEP 2:  Statistical Analysis

#### - Assumptions and Stats 
```{r treatment }
multi_results <- multi_flow_test(
  data = filtered_data,
  measurements = measurements,
  group_var = "treatment",
  subject_id = "exp_no",
  stratify_var = "genotype",       # compare treatments within each genotype
  day_var = "day",
  value_var = "value",
  save_dir = save_path,
  comparison = paste0(exp_name, "_", comparision)
)

saveRDS(multi_results, file = file.path(rdata_path, paste0(exp_name, "_", comparision, "_stat_results.rds")))
```



## STEP 3 : PLOTS with stat annotation

```{r subfolder }
plots_folder <- file.path(save_path, "Plots")
make_subdir(plots_folder)
```

##### - Generate plots with annotaions
```{r}
filtered_summary <- filtered_summary %>% filter(day != 3)
filtered_raw <- filtered_data %>% filter(day != 3)

plot_list <- list()

for (measure in measurements$measurement) {
  for (geno in unique(filtered_data$genotype)) {
    geno <- as.character(geno)
    # Create the plot
    p <- make_flow_plot(
      measurement_name = measure,
      summary_data = filtered_summary,
      raw_data = filtered_raw,
      cell_val = geno,  
      folder = save_path,
      exp_name = exp_name,
      color_map = color_map
    )
    
        # 🪄 Override guides based on cell type
    if (grepl("D21", geno)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
          color = "none", 
          alpha = "none"
        )
    } else if (grepl("T21", geno)) {
      p <- p + 
        scale_fill_manual(name = NULL, values = setNames(color_map$Color, color_map$Group))+
        guides(
          fill = "none",
          color = guide_legend(order = 1, override.aes = list(fill = NA)),
          alpha = "none"
        )
    }

    # Annotate stats 
    p_annot <- add_multi_annotation(
      plot = p,
      stats_data = multi_results,           
      measurement = measure,
      group_col = "treatment",                
      stratify_by = "genotype",             
      stratify_level = geno,
      show_effect = FALSE,
      show_pval = FALSE
    )
    
    p_annot <- add_friedman_annotation(
      plot = p_annot,
      stats_data = multi_results,
      measurement = measure,
      stratify_by = "genotype",
      stratify_level = geno
    )

    # Save it
    ggsave(filename = paste0(exp_name, "_", measure, "_", geno, "_stats.pdf"),
      plot = p_annot, path = plots_folder,
      width = 6, height = 7, units = "in"
    )
    
    plot_key <- paste0(measure, "_", geno)
    plot_list[[plot_key]] <- p_annot
  }
}

saveRDS(plot_list, file = file.path(rdata_path, paste0(exp_name, "_alltreatments_plots.rds")))


```

